\documentclass[11pt]{article}
    \title{\textbf{Default Doc}}

    \date{\the\day/\the\month/\the\year}

\input{../pdf.tex}
\input{../webmacros.tex}
\input{../programmingmacros.tex}
\input{../amsthmstuff.tex}

\usepackage{amsmath}
\begin{document}

% Review and Introduction to Iterative Techniques for $Ax=b$ Solvers

I mentioned in a \hyperlink{https://clockssugars.blog/articles/0126-heateq/}{previous post} that I was considering making my own finite element solver. I had already done much of the necessary reading for how one does this (a topic I should discuss in a separate post), which I then had to buttress with some additional double checking of methods on basis functions and boundary conditions, etc. etc. Anyway, the thing that got me stuck was that finite element techniques owe much of their stability and solvability to statements that can be made about the resulting discretized system as a linear relation, something of the form $Ax = b$. The fast version of this is that information from the boundary conditions (or perhaps state iteration in the case of a time-dependent differential equation) in addition to conditions on derivatives, inform a notion of error or deviation from satisfaction of the diff eq. which, when linear, can solve for a minimum-error state. (When not linear, other techniques may be used.)

The trouble is that these linear systems end up huge. For a first-order linear system (i.e. we describe our solution function in terms of linear functions, or first order polynomials) we have something like one basis function per vertex of our space discretization graph. Since we usually do this with triangles in 2D, a 20 by 20 grid ends up with a linear system of dimension in the hundreds. Some systems require higher order basis functions in order to derive stable solutions, e.g. for second order, each vertex and each edge are associated to a second order basis function (in fact this is similar to subdividing each triangle into four, then once again using vertices for basis functions), meaning a roughly fourfold increase in the number of basis functions. The most obvious system that suffers from instability without second order basis functions is Navier-Stokes systems which involve not just a scalar field solution, but a vector field (the field describing fluid flow), meaning another factor of three increase to the number of basis functions. In total, you could end up with a $Ax=b$ system of dimension in the thousands or tens of thousands that must be solved on every timestep of the simulation before any interesting phenomena are observed, and this is before methods to improve stability such as Runge-Kutta which may require doing the calculation two or more times (usually four) per time step.

Put this way, such a simulation seems like a ridiculous computational task, even though mathematically it is not all too complicated (in my opinion). There is a trick here however, that the matrix $A$ we come up with for our $Ax=b$ problem is quite sparse. Despite potentially tens of thousands of dimensions of the overall problem, each dimension, associated to a basis function which is itself associated to a vertex or edge, is generally only coupled to its neighboring vertices and edges, meaning almost all of the $A$ matrix is just zeros. This information on its own is useless if we are naive about how to resolve a $Ax=b$ system; methods such as Gaussian elimination or $LU$ decomposition will rapidly couple every dimension to every other in the course of narrowing down a single solution. Moreover, since we are working with floating point arithmetic, the error due to rounding may ultimately be significant, in which case methods which spend significant time to find an exact solution to $Ax=b$ may be inferior to methods which can rapidly approximate a solution to within a tolerable amount of error.

Following my previous post, discussing a method to solve the heat equation on mobile devices via webgpu, I had assumed there would be some way to parallelize the solving of an $Ax=b$ problem via mere GPU distribution of the calculation. Of course, this was a bit of a handwave on my part, and it quickly became apparent that most other finite element systems simply used an existing library for numerical solutions of large linear systems, such as \hyperlink{https://en.wikipedia.org/wiki/Basic_Linear_Algebra_Subprograms}{BLAS} or \hyperlink{https://en.wikipedia.org/wiki/Portable,_Extensible_Toolkit_for_Scientific_Computation}{PETSc}. These would not be available to me if I wanted my finite element solver to run on a mobile device, where my methods would need to be implemented in WASM and WebGPU, and so I would need to learn some details of the implementation myself.

The following documents my exploration of this topic, of iterative methods for solving large $Ax=b$ systems in the interest of finite element methods (esp. for Navier-Stokes), and in particular, documents my reading of Yousef Saad's \emph{Iterative Methods for Sparse Linear Systems} (2003) and Gerard Meurant's \emph{Computer Solution of Large Linear Systems} (1999). As I write this I can hardly say that I have read either book thoroughly or that I have a deep grasp of the topics within; indeed, I had quite some trouble placing the intuitions these authors appear to have intended within any context I am familiar with. I write this in part to straighten out my own thoughts as I attempt to understand the volumes, as well as to make explicit a narrative I am forming about how one understands the topic.

Thus, this article endeavors to serve as a reference for readers with some linear algebra knowledge who wish to quickly develop some understanding of the field, and as a reference for myself in future, should I forget some of what I presently know.

\subsection{Nothing Ventured, Nothing Gained: What Naive Investigation Will Get You}

In my first exploration of the techniques available to beat methods for exact solution of $Ax=b$ problems, I repeatedly found that either an algorithm either appeared to be a method for exact solution in a trench coat, or the reason why it was better than that could only be expressed through some handwaving. As I expect readers curious about these topics are likely to come to similar conclusions initially, we'll first outline these pitfalls.

The shape I will paint these pitfalls is that which they appeared to me, which is to say that they will appear as $QR$-decomposition. That is, if we take $A$ and perform the Gram-Schmidt algorithm on its columns, we will obtain an orthonormal basis, the columns of which form $Q$, as well as the coefficients required to put the columns in $Q$ back together into $A$, forming the matrix $R$. Since the columns or $Q$ are orthonormal, it is an orthogonal matrix, and since the coefficients in the Gram-Schmidt algorithm are obtained by successively subtracting the components of previously found orthonormal vectors, the $n$-th column of $R$ can only contain $n$ coefficients, so we are speaking of an upper-triangular matrix. This decomposition is then written $A = QR$ since the matrix product reconstructs the original matrix $A$.

Completing such an operation on $A$ is half the job of solving $Ax=b$ directly; orthogonal matrices such as $Q$ have the property that $Q^T = Q^\inv$, meaning that we need only flip $Q$ to find its inverse, and triangular matrices are sought after precisely because once $c = Q^T b$ is computed, the equivalent system $Rx = c$ is especially easy to solve by back substitution: since $R$ is upper triangular, solve $R_{nn}x_n = c_n$ (i.e. $Rx = c$ in it's $n$-th dimension) since the $n$-th row of $R$ is all zero except for the final element. With $x_n$ solved, use that to solve for $x_{n-1}$ since \begin{gather*} R_{(n-1) n} x_{n} + R_{(n-1) (n-1)} x_{n-1} = c_{n-1} \end{gather*}
by the triangular shape of $R$. So on and so forth, at any given time there will only be one unknown.

But both in the $QR$-decomposition step and in the $Rx = c$ back-substitution step, we are dealing with an algorithm that is roughly $O(n^2)$. Moreover, we take no advantage of the sparsity of $A$ that is previously mentioned, since the Gram-Schmidt algorithm must necessarily couple all dimensions together; our $Q$ and $R$ matrices will be anything but sparse, and so we surrender whatever advantages we had.

This is the technique we are trying to beat, and without deeper thought, it will at first seem like nothing we do escapes this.

\subsubsection{Conjugate Gradient: A First approach}

We cannot begin this exploration (and indeed we will be unable to end it) without discussing the conjugate gradient method. When I first began investigation of iterative methods for solving linear systems, conjugate gradient was the name that everyone brought up, which lead me to the wikipedia page and the corresponding context-agnostic understanding. We will find in later sections that conjugate gradient is more of a family of techniques (and an extremely effective family at that) but you certainly wouldn't guess that from a quick reading of given on wikipedia, so we'll start with what is described there.

The wikipedia page describes the algorithm as requiring $A$ to be symmetric ($A = A^T$) and positive definite ($x^TAx > 0$ for all $x \in \reals^n$) thus also an invertible matrix. These conditions together are actually also the conditions for an inner product, so we have in effect defined an inner product $\langle u, v \rangle_A = u^T A v$. With this inner product in hand, we now require a $A$-orthogonal basis $\vec{p}_i$ such that $\vec{p}_i \neq \vec{0}$ for all $i = 1,...,n$ and $\langle \vec{p}_i, \vec{p}_j\rangle_A = 0$ if $i \neq j$.

Since we are describing an iterative method, what we will do is start with some $\vec x_0$ and use it to define each successive improvement $\vec x_{j+1}$ until hopefully our $\vec{x}_j$ is either an exact solution $\vec x_*$ satisfying $A\vec x_* = b$ or is close enough. In the conjugate gradient method, these successive improvements at each step $j$ are defined by subtracting the $\vec{p}_j$ component of the error. The idea is as follows:

We desire $A\vec{x}_* = b$, and so $A\vec{x}_j - b = \vec{r}_j$ will be a non-zero residual vector whenever $\vec{x}_j \neq \vec{x}_*$ (or indeed if they are equal, terminate and deliver solution $\vec{x}_*$). Our update step, in removing the $\vec{p}_j$ component of the error, will be \begin{gather*}
\vec{x}_{j+1} = \vec{x}_j + \alpha_j \vec p_j
\end{gather*}
where $\alpha_j$ is the amount of error in the $\vec{p}_j$ direction. If we knew $\alpha_j$, we would propogate to the next step and find \begin{align*}
\vec r_{j+1} &= A(\vec x_j + \alpha_j \vec p_j) - b \\
&= A\vec x_j + \alpha_j A\vec p_j - b \\
&= \vec r_j + \alpha_j A\vec p_j
\end{align*}
where, by $A$-orthogonality of $\vec{p}_j$ vectors, the correction term $\alpha_j A \vec p_j$ would be orthogonal to all other $\vec{p}_k$ where $k \neq j$.

Now at this stage, we have not promised that our set of $\vec{p}_j$ vectors are \emph{normal} vectors, so they may have arbitrary magnitude; this means that our calculation of $\alpha_j$ is downstream from our choice of $\vec{p}_j$. Since ultimately we want our $\vec{r}_j$ vector to become the zero vector, we should choose our $\vec{p}_j$ so that each $\alpha_j A \vec{p}_j$ can subtract as much of itself as is present from the $\vec{r}_j$ it acts on. This implies the following \begin{gather*}
\alpha_j = \frac{\langle \vec p_j, \vec r_j \rangle}{\langle \vec p_j, \vec p_j \rangle _A } = \frac{\vec p_j \cdot \vec r_j }{\vec p_j \cdot A \vec p_j } \\
\vec p_j = \vec r_j - \sum_{i < j} \frac{\vec r_j \cdot A \vec p_i}{\vec p_i \cdot A \vec p_i} \vec{p_j}
\end{gather*}
which is to say, we choose each $\vec p_j$ basically by running Gram-Schmidt on each residual as we get it, and we calibrate the $\alpha_j$ to subtract from $\vec x_{j}$ what will become the error in $\vec r_j$.

If you don't understand this at this point, that's okay. Neither did I. At this point the wikipedia page provides an algorithm in which there appears some way to compute each $\vec p_j$ in the basis in $O(n)$ time besides the residuals as they are calculated, but the page does not particularly explain how other than by reference vague to Krylov subspaces and orthogonality of $\vec r_j$ vectors. Absent that explanation, it is also easy to see the Gram-Schmidt calculation of the $\vec p_j$ vectors and conclude that we are still probably not making use of the sparsity of $A$.

\figuresvgwithcaption{./Conjugate_gradient_illustration}{(Credit to Wikipedia, from the page on the Conjugate Gradient method) This illustration from the wikipedia page seems to give the intuition that the Conjugate Gradient algorithm merely picks a series of directions and iteratively subtracts their components from the error as though it were a circumstance specific improvement on Gradient Descent. Moreover, this image may lead you to the intuition that Conjugate Gradient must proceed through all $n$ iterations to avoid risk of a radically incorrect answer, which we will see later is not true.}

With the Gram-Schmidt in the picture, it appears for now as though all Conjugate Gradient has offered us is a faster way to do $QR$-decomposition. My frustration with this lead me to read deeper.

\subsubsection{Preconditioners: A First Approach}

Similarly, the concept of preconditioners had been mentioned to me, both by friends and appearing in documentation related to existing implementations of finite element solvers. Morally, the idea as I initially understood it, was that a preconditioner matrix $M$ may exist such that $M^\inv$ is easy to calculate, and where $M^\inv A x = M^\inv b$ would be easier to solve than $Ax = b$ directly, even though it would still solve $Ax=b$.

More generally, a preconditioner aims to reduce the \textbf{condition number} of the problem, whcih we can think of as how sensitive the residual is to an error in $x$. Recall from earlier that for the application of finite element problems, $Ax=b$ implies minimum possible violation of some differential condition and $Ax - b \neq \vec 0$ implies greater violation of that condition. We would then say that if $A$ is such that a very small error in $x$ away from the $\vec x_*$ that solves $A\vec x_* - b = \vec 0 $ can lead to a very very significant violation in the differentiability condition we want, we say that $A$ is \textbf{ill-conditioned}. Now, we can't do anything about the differentiability condition itself (although we could perhaps pick a $Bz = b$ problem to solve where $Bz = Ax$ via $Cz = x$ and $BC^\inv = A$, changing the basis we solve in) but if we can choose a good preconditioner $M$, then we may well solve $M^\inv A x = M^\inv b$ faster, or eliminate significant sources of error in a candidate $x$ sooner.

The formal way to compute a condition number is actually to assume we find an exact solution $x$ but there is some error in the $b$ for which we use $Ax = b$ to calculate $x$. Posed that way, we define it as \begin{align*}
\kappa(A) :&= \max_{e,b \neq \vec 0} \left\{ \frac{\lVert A^\inv e \rVert}{\lVert A^\inv b \rVert} / \frac{\lVert e \rVert }{\lVert b \rVert} \right\} \\
&= \max_{e \neq \vec 0} \left\{ \frac{\lVert A^\inv e \rVert}{\lVert e \rVert} \right\} \max_{b \neq \vec 0} \left\{ \frac{\lVert b \rVert }{\lVert A^\inv b \rVert } \right\} \\
&= \max_{e \neq \vec 0} \left\{ \frac{\lVert A^\inv e \rVert}{\lVert e \rVert} \right\} \max_{b \neq \vec 0} \left\{ \frac{\lVert A x \rVert }{\lVert x \rVert } \right\}
\end{align*}
for $e$ error in $b$. The normal way to express this would be to use an operator norm, writing $\kappa(A) = \lVert A^\inv \rVert \lVert A \rVert$ but it is just as well to leave it in this form, since a larger population of people can read this expression and correctly conclude that it roughly means $\kappa(A)$ is usually the product of $A$'s largest eigenvalue squared with the inverse of its smallest eigenvalue squared. That is, each maximum factor is $\sqrt{Ax \cdot  A x}$ divided by $\sqrt{x \cdot x}$, so normalized but with $x$ chosen to extract the maximum increase from $A$ which is only done if $x$ is the eigenvector with greatest eigenvalue. There are exceptions to this intuition (in general it's more like the square root of the greatest eigenvalue of $A^T A$), but presuming $A$ is diagonalizable, what we are saying is $A$ is worse conditioned when there is greater disparity between the magnitude of its eigenvalues.

If you're anything like me, upon hearing this you may immediately feel that we have discussed a method that might help us reducing the disparity between the magnitude of eigenvalues. In fact, an orthogonal matrix as earlier mentioned has only eigenvalues of magnitude one, so a perfect preconditioner $M$ that achieves preconditioning to minimum condition number would be such that $M^\inv A = Q$ is an orthogonal matrix, and that's just $QR$-decomposition again isn't it. Yeah. Yeah we were just about to deduce the perfect $M$ is $M=L$ where $L$ is a lower triangular matrix deduced by applying $QR$-decomposition to $A^T = QL^T$ (with $L^T=R$ upper triangular) so that $A = L Q^T$, in which case the preconditioner $M = L$ results in $M^\inv A = L^\inv (LQ^T) = Q^T$ which is orthogonal and thus perfectly preconditioned.

So once again, nothing is gained from considering preconditioning naively like that. We find ourselves falling back on previously known naive methods.

\subsection{Viable Frameworks and Some Signposting}

What I learned quickly when reading the two books as well as investigating other sources was that most writers on iterative techniques have a lens from which they view techniques, and generally handwave details that don't fit neatly into their lens. Moreover, the books repeatedly closed in on generalized frameworks of methods for which each technique seemed merely like a choice of certain parameters.

What I'll try to do here is focus on intuitive explanations but not single any of them out as canonical. Different systems of thinking about these techniques reveal different things, and some of them are better suited to make obvious why a thing is true or why an algorithm is efficient. Worse, the implications of each way of thinking is generally necessary to understand the nuances of the others.

We will proceed by first discussing splitting techniques, then use the framing of projection techniques, showing in our exposition that they are ultimately different ways of thinking about the same thing. With those two frameworks apparent, we will proceed to Krylov subspaces, where the core insight of the conjugate gradient technique will become apparent, as well as how we can manipulate this insight to form alternative algorithms purpose-fit to other situations.

\subsection{Splitting Methods}

Consider the component-wise statement of what we want. $Ax_*=b$ with $A$ containing elements $A_{ij}$ at row $i$ and column $j$ and $x_*$ and $b$ containing elements $x^j_*$ and $b^i$ (we use superscripts to mean index on vectors since we will use subscripts to indicate different vectors; matrices still use subscripts for indices) respectively, we want \begin{gather*}
b^i - A_{ij} x^j_* = 0 
\end{gather*}
for all $1 \le i \le n$. If we have some decomposition $A = M - N$ where we describe $A$ as the difference of two matrices (the choice of to reuse $M$ is not an accident here) then this requirement can be written as follows: \begin{gather*}
b^i + N_{ij} x^{j}_* = M_{ij} x^{j}_*.
\end{gather*}
Now, with $x_*$ as a solution to $Ax=b$, this holds true on its own. But if $x_0$ is merely an initial guess vector, then we can define a recurrence relation \begin{gather*}
b^i + N_{ij} x^j_k = M_{ij} x^j_{k+1}
\end{gather*}
which is given explicitly by \begin{gather*}
x_{k+1} = M^\inv b + M^\inv N x_k
\end{gather*}
assuming we chose our splitting $A = M - N$ in such a way that $M$ is easily inverted, such as by choosing $M$ as a diagonal matrix. Now, you may be asking as I did, this is a nice recurrence relation but there is no reason to think yet that it necessarily brings us closer to solving for $x_*$. For that, we must choose $M$ and $N$ very judiciously, and our choice will have consequences in all sorts of ways, for the computing complexity of our algorithm and for the reusability of products of the algorithm.

In general, any iterative method of the form \begin{gather*}
x_{k+1} = G x_k + f
\end{gather*}
will converge so long as the \textbf{spectral radius} of $G$ is less than one. \begin{definition}[Spectral Radius]
Let $A$ be a $n \times n$ matrix with eigenvalues $\lambda_1, \dots, \lambda_n$. Then we define $\rho(A)$ the \textbf{spectral radius} to be $\rho(A) = \max \{ |\lambda_1|, \dots, |\lambda_n| \}$ the largest magnitude eigenvalue.  
\end{definition}
The spectral radius coincides with the operator norm described above as $\max_{x \neq 0} \lVert Ax \rVert / \lVert x \rVert$ for \emph{normal matrices} i.e. matrices which satisfy $A^\dagger A = A A^\dagger$, which for real matrices is just $A^T A = A A^T$ (and it's easy to see why such a condition would lead to the square root of eigenvalues of $AA^T$ coinciding with eigenvalues of $A$). The operator norm has its uses when $A$ is not square since an $n\times m$ has $A A^T$ a $n \times n$ shape, or $A^T A$ with a $m \times m$ shape.

\begin{proposition}[Convergence of Iterative Methods]
An iterative method $x_{k+1} = Gx_k + f$ with $x_{k},f \in \reals^n$ and $G \in \reals^{n \times n}$ converges if $\rho(G) < 1$.
\end{proposition}

\begin{my proof}{m}
If $G$ has $\rho(G) < 1$ then we know not only that the largest magnitude eigenvalue is less than one but that, since it is the largest magnitude, all eigenvalues of $G$ have magnitude less than one. Recall that eigenvalues are those $\lambda \in \reals$ such that \begin{gather*}
\det(G - \lambda I) = 0
\end{gather*}
since eigenvectors must satisfy $Gv = \lambda v = \lambda I v$ and thus $Gv - \lambda I v = \vec 0$, placing eigenvectors in the null-space of $(G - \lambda I)$, and making the determinant zero since it has non-trivial null-space.

Since we have guarenteed all eigenvalues of $G$ are $|\lambda| < 1$, we have also guarenteed that $\lambda \neq \pm 1$ for all eigenvalues $\lambda$, and so we know \begin{gather*}
\det(G - I) \neq 0 \\
\det(G - (-1)I) \neq 0 
\end{gather*}
merely because $1$ and $-1$ are known not to be eigenvalues due to $\rho(G) < 1$.

We conclude from this in particular that $G - I$ is an invertible matrix, and thus $I - G$ is an invertible matrix.

Using this information, return to our recurrence relation. To ask if \begin{gather*}
x_{k+1} = G x_k + f
\end{gather*}
converges is to ask if there exists $x_* \in \reals^n$ such that $x_* = G x_* + f$. Such a condition can also be written \begin{align*}
x_* - G x_* & = f\\
(I - G) x_* &= f.
\end{align*}
Since we have deduced $I - G$ is invertible however, we know there exists $x_* = (I - G)^\inv f$ as a fixed point of this iteration.

Now we just need to know if we will actually converge towards this fixed point. For that, assume we have an initial candidate vector $x_0 \in \reals^n$ and express it in terms of the fixed point and some error away from the fixed point $x_0 = x_* + \varepsilon$ with $\varepsilon \in \reals^n$. With this decomposition, our recurrence relation appears as \begin{align*}
x_{1} &= Gx_0 + f \\
&= G(x_* + \varepsilon) + f\\
&= Gx_* + f + G \varepsilon \\
&= x_* + G \varepsilon
\end{align*}
If we decompose our sequence of vectors $x_{k}$ defined by the recurrence relation starting from some $x_0$ as $x_k = x_* + \varepsilon_k$ then we can see clearly by the above reasoning that \begin{gather*}
x_{k+1} = x_* + G \varepsilon_k
\end{gather*}
and so the difference between any two steps $x_{k+1}$ and $x_k$ must be \begin{align*}
x_{k+1} - x_k &= (x_* + G \varepsilon_k) - (x_* + \varepsilon_k) \\
&= G \varepsilon_k - \varepsilon_k.
\end{align*}
But by $x_{k+1} = x_* + \varepsilon_{k+1}$ we can also see \begin{align*}
x_{k+1} - x_k &= (x_* + \varepsilon_{k+1}) - (x_* + \varepsilon_k) \\
&= \varepsilon_{k+1} - \varepsilon_k
\end{align*}
so together, we conclude a recurrence relation on the error components of $x_k$, \begin{gather*}
G \varepsilon_k - \varepsilon_k = \varepsilon_{k+1} - \varepsilon_k \\
G \varepsilon_k = \varepsilon_{k+1}
\end{gather*}
meaning $x_k = x_* + G^k \varepsilon_0$. We must now show this error term converges to zero, i.e. that $G^k$ in the limit goes to the zero matrix.

Take the Jordan normal form of $G$ (that is, we require a decomposition $G = PDP^\inv $ but do not wish to require $G$ is diagonalizable, so $G = P J P^\inv$ with $J = D + N$ with $D$ diagonal and $N$ nilpotent will sufffice). The matrix $D$ will contain the eigenvalues of $G$ along its diagonal (possibly with repeats) which we know from earlier to all be $|\lambda| < 1$. By cancelling inverses of $P$ we have the standard technique \begin{gather*}
G^k = (PJP^\inv) ... (PJP^\inv) = PJ^k P^\inv
\end{gather*}
Recall nilpotent matrices are defined to have the property that some $p \in \nats$ satisfies $N^p = \mathbf{0}$ the zero matrix, and jordan normal form results in $PJP^\inv$ with $J = D + N$ with $N$ nilpotent. If we look at the polynomial expansion (expressed combinatorically) of $J^k$, \begin{gather*}
J^k = (D + N)^k = \sum_{i=0}^k \left( \begin{matrix} k \\ i\end{matrix} \right) D^{k - i} N^i
\end{gather*}
then we know by the property that $N^p = \mathbf{0}$ that only the first $p-1$ terms will be non-zero. So assuming $k \ge p$, we have 
\begin{align*}
J^k &= \sum_{i=0}^{p-1} \left( \begin{matrix} k \\ i\end{matrix} \right) D^{k - i} N^i \\
&= D^k \left( \sum_{i=0}^{p-1} \left( \begin{matrix} k \\ i\end{matrix} \right) D^{- i} N^i \right)
\end{align*}
Now, $D^k$ should tend towards zero exponentially since it contains the eigenvalues of $G$ along its diagonal, and these are all known to be $|\lambda| < 1$, but we need to make sure that it will tend towards zero faster than the up-to $p-1$ combinations of $k$ factor will grow as we increase $k$. For this, recall that \begin{gather*}
\left( \begin{matrix} k \\ i\end{matrix} \right) = \frac{k!}{i! (k-i)!}
\end{gather*}
meaning that if we fix $i$, say at $i=2$ for example, we find \begin{gather*}
\left( \begin{matrix} k \\ 2 \end{matrix} \right) = \frac{(k-2)!(k-1)k}{2! (k-2)!} = \frac{1}{2!} (k-1)k
\end{gather*}
so it is clear that fixing the number of combinations $i$ produces a polynomial of order $i$. Knowing this, the fastest growing term in the summation above should be the term for which $i=p-1$. For large $k$ this term will dominate, so let us write \begin{align*}
J^k &\approx D^{k} \left( \prod_{i=0}^{p-1} (k-i) \right) D^{-p+1} N^{p-1} \\
&\approx D^{k} \left( \sum_{i=0}^{p-1} \left( \begin{matrix} p-1 \\ i \end{matrix} \right) k^i \right) D^{-p+1} N^{p-1}
\end{align*}
where in this form it is clear that the exponential decay of $D^k$ will go to zero faster than any polynomial in $k$. We could also have made this argument prior to the approximation we made, but it is clear in any case that finite order polynomials will grow slower than an exponential can go to zero in the limit.

We conclude that $G^k$ must go to zero as $k \to \infty$ and thus $\varepsilon_k \to \vec 0$ as $k \to \infty$, so any sequence $x_k$ with the described recurrence relation has $x_k \to x_*$.
\end{my proof}

This proof also can be read to imply that the smaller $\rho(G)$ is, the faster the method should converge.

So the pattern with a splitting method will be thus: pick a splitting $A = M - N$, make sure it results in $\rho(M^\inv N) < 1$, and make sure we can iterate our recurrence relation cheaply.

\subsubsection{Jacobi Iteration}

One very simple choice, as alluded to, is simply to consider a decomposition $A = D + L + U$ where $D$ is simply the elements of $A$ on the diagonal, $L$ is those below the diagonal, and $U$ is above the diagonal, and choose $M = D$ and $N = - L - U$. \begin{gather*}
L_{ij} = \left\{ \begin{matrix} A_{ij}, & i < j \\ 0 & \text{else} \end{matrix}\right. \hspace{3em}
U_{ij} = \left\{ \begin{matrix} A_{ij}, & i > j \\ 0 & \text{else} \end{matrix}\right. \\
D_{ij} = \left\{ \begin{matrix} A_{ij}, & i = j \\ 0 & \text{else} \end{matrix}\right.
\end{gather*}

Jacobi iteration can be known to converge if $A$ is \emph{strictly diagonally dominant} i.e. \begin{gather*}
|A_{ii}| > \sum_{j \neq i} |A_{ij}|
\end{gather*}
that is, the element on the diagonal is larger than the sum of all other elements on its row. We can suffice with weakening $>$ to $\ge$ if we also know that the matrix is also \emph{irreducible}. This condition is a little more complicated but can be thought of roughly as whether the matrix can be described as two mostly-separate systems. For instance, one definition of \emph{reducible} matrices $A$ is that there exists a permutation matrix $P$ such that $P^\inv A P$ has block structure \begin{gather*}
P^\inv A P = \begin{bmatrix} D_1 & \mathbf{0} \\ F & D_2 \end{bmatrix}
\end{gather*}
where $F$ is a square matrix, $\mathbf 0$ the zero matrix, and $D_1,D_2$ diagonal matrices. Another equivalent condition of matrix reducibility is that the index set $I = \{ 1, \dots , n \}$ has a partition into $I_1, I_2$ (partition meaning $I_1 \cap I_2 = \emptyset$ and $I_1 \cup I_2 = I$, they are disjoint and union to the full set) has some pair $i \in I_1$ and $j \in I_2$ with $A_{ij} = 0$. When $A$ is \emph{irreducible} (for all such partitions, every $i\in I_1,j \in _I2$ has $A_{ij} \neq 0$) and it is diagonally dominant (not necessarily strictly) then besides knowing convergence due to $\rho(M^\inv N) = \rho(- D^\inv (L + U)) < 1$, we also know the technique converges with $A$ strictly diagonally dominant or irreducibly diagonally dominant without finding $M^\inv N$'s largest eigenvalue.

Proofs of these convergence conditions are present in chapter 5.2 of Gerard Meurant's book.

\subsubsection{The Gauss-Seidel Method}

The Gauss-Seidel method consists of a splitting method $A = M - N$ where with the decomposition $A = D + L + U$ into a diagonal component $D$, a component beneath the diagonal $L$, and a component above the diagonal $U$, we set $M = D + L$ and $N = -U$.

You may immediately look at this and think \dquote{hold on, we know that $D$ is easy to calculate the inverse of, its just a diagonal matrix, but we have no guarentee of this for $L$} and that is true. This is an interesting situation where we need to declare that matrices and linear algebra actually form a poor abstraction.

Consider the elementwise statement of Jacobi iteration: \begin{gather*}
x_{k+1}^i = \frac{1}{A_{ii}} \left( b^i - \sum_{j \neq i} A_{ij} x_k^j \right).
\end{gather*}
This is because $D$ is just $A$'s diagonal elements so multiplying by $D^\inv$ is multiplying each $i$'th element by $1/A_{ii}$, and the remaining multiplication of $x_k$ by $N = L + U$ is just $A$ without its diagonal elements; we do a standard matrix multiplication but skip when $j=i$.

Notice that with this iteration, we commit to calculating every element of $x_{k+1}$ before moving on to $x_{k+2}$. But it may well be the case that part way through calculating $x_{k+1}$, we will have removed enough error from the earlier elements of the vector that we could get some efficiency gains by using the new elements early. If we were to do this, our iteration step would instead look like \begin{gather*}
x_{k+1}^i = \frac{1}{A_{ii}} \left( b^i - \sum_{j > i} A_{ij} x_k^j - \sum_{j < i} A_{ij} x_{k+1}^j \right).
\end{gather*}
Stated as matrices, this is \begin{gather*}
D x_{k+1} = b - Ux_k - Lx_{k+1}
\end{gather*}
and thus \begin{gather*}
(D +  L)x_{k+1} = b - Ux_k
\end{gather*}
making clear that this method does indeed describe $M = D + L$. Naively proceeding from the matrix abstraction, we would appear to have a method which requires the costly calculation of $(D + L)^\inv$, but as above, this is not the case!

What appears as a costly calculation is merely the side effect of expressing a technique which weakens the assumptions of linear algebra (namely the simultaneity of vector operations) in the terms of linear algebra. As long as we calculate each element of $x_{k+1}^i$ in order (ascending $i$), the only parts of $x_{k+1}$ we ever use are those we have already calculated, so no matrix inversion is necessary!

This forms a lesson that will be important to remember moving forward: a non-diagonal choice of matrix $M$ does not necessarily mean an inversion is necessary \emph{if} we are clever about its non-diagonal components and our iteration algorithm. In fact, the insistence at \emph{every} turn that what we are doing here is linear algebra will lead us astray; what we are doing is coming up with algorithms, and linear algebra does not have the final say.

Gauss-Seidel is similarly known to converge if $A$ is strictly diagonally dominant.

\subsubsection{Successive Over Relaxation}

This sequence of techniques would not be complete if I did not mention Successive Over Relaxation (SOR) although I must say it is the technique I spent the least time thinking about. Superficially, one interprets the SOR method as the choice of some $\omega > 0$ and a splitting \begin{gather*}
A = D + L + U \\
A = M - N \\
M = \frac{1}{\omega}D + L \\
N = \frac{1-\omega}{\omega} D - U 
\end{gather*}
Written as an matrix iteration, it typically appears as \begin{gather*}
(D + \omega L) x_{k+1} = \omega b  - \omega U x_k + (1 - \omega) D x_k
\end{gather*}
but as a component-wise iteration it appears more obviously as a Gauss-Seidel style mid-iteration update method \begin{gather*}
x_{k+1}^ i = \frac{\omega}{A_{ii}} \left(b_i - \sum_{j < i} A_{ij} x^j_{k+1} - \sum_{j > i} A_{ij} x^j_{k} \right) + (1 - \omega) x_k^i.
\end{gather*}
That is to say, what we do is roughly to take linear combinations of what Gauss-Seidel tells us to do and our previous timestep, perhaps to provide a measure of stability. I say roughly because we by no means confine $\omega$ to the interval $(0,1]$. In fact, the result referred to here is that \dquote{If SOR converges then $0 < \omega < 2$} so we may also accelerate the alterations Gauss-Seidel suggests. The idea of extending $\omega$ to the range $(0,2)$ provides the prospect that we may have some parametric control over the \emph{convergence rate} of the technique, i.e. there may be an ideal $\omega$ for which the technique converges \emph{fastest}.

Now, the statement that SOR converges implies $\omega \in (0,2)$ does not mean in general that $\omega \in (0,2)$ guarentees convergence of SOR; in chapter 5.4 Meurant shows that SOR converges if under the condition that $A$ is a \squote{H-matrix} and its associated $A = D + L + U$ has \begin{gather*}
0 < \omega < \frac{2}{1 + \rho(|D^\inv (L + U)|)}.
\end{gather*}
Now, what an \squote{H-matrix} is is an irritating conversation that I prefer to largely skip over here. A definition is provided for completeness. \begin{definition}
Let $A \in \reals^{n \times n}$ be a matrix. We say that $A$ is a \textbf{M-matrix} if there exists $\sigma \in \reals$ and $B \in \reals^{n\times n}$ such that \begin{gather*}
A = \sigma I - B,\\
\sigma > 0, \\
\forall i,j \ \ B_{ij} \ge 0, \\
\rho(B) \le \sigma.
\end{gather*}
Now associate to $A$ a matrix $M^A \in \reals^{n \times n}$ which is defined \begin{gather*}
M^A_{ij} = \left\{ \begin{matrix} |A_{ij}| & i = j \\ -|A_{ij}| & i \neq j \end{matrix} \right.
\end{gather*}
We say that $A$ is an \textbf{H-matrix} if $M^A$ is an M-matrix.
\end{definition}

All over Meurant's book there is frequent reference to the H-matrix condition being a requirement for convergence of techniques. As we will see later, we will have even better techniques that do not require such an unwieldy property.

As for the optimum $\omega$, Meurant cites a 1950 Ph.D. thesis by a David M. Young Jr who appears also to have (simultaneously) invented the SOR method. Assuming certain hypotheses, the optimum $\omega$ for a $A x = b$ problem with $A = D + L + R$ is defined \begin{gather*}
\omega_b = \frac{2}{1+\sqrt{1-\rho(-D^\inv(L+U) )^2 }}
\end{gather*}
although Meurant mentions that having to calculate $\rho(D^\inv(L + R))$ makes this formula impractical most of the time, and so there exist a variety of techniques to approximate it.

\subsection{Projection Methods}

Much in the way that one might look at the initial statement of splitting methods, a priori, and say \dquote{on its own, there's no reason that should do anything useful} and then add certain assumptions to ensure that a method converges, projection methods lay out only \squote{an operation} with no guarentees attached. In fact even less than \squote{an operation}, projection methods call for defining two subspaces of $\reals^n$ called $\mathcal{K}$ and $\mathcal{L}$ and saying, given a vector $x_0 \in \reals^{n}$, we want to calculate a \emph{some} vector $x_1 \in \reals^{n}$ which is $x_1 \in \mathcal{K}$ and $ b - Ax_1 \perp \mathcal{L}$ i.e. the residual $r_1 = b - Ax_1$ is entirely perpendicular to the space $\mathcal{L}$. Now, two asterisks must follow. First, it is often the case that what we calculate is not as the \squote{next in the sequence} $x_1$ but the difference between approximations, i.e. we do not calculate $x_1$ with these properties, we calculate $\delta_0 \in \mathcal{K}$ and $r_0 - A\delta_0 \perp \mathcal L$ and set $x_1 = x_0 + \delta_0$. \begin{align*}
\text{(Note:)} \hspace{1.5em} \mathcal{L} \perp &\hspace{0.6em} b - Ax_1 \\
&= b - A(x_0 - \delta_0) \\
&= (b - Ax_0) - A\delta_0 \\
&= r_0 - A \delta_0.
\end{align*}
Second, this method also forms an iteration technique so while the notation does not appear to be common, it is perhaps more appropriate to think of these choices of subspaces as $\mathcal{K}_k$ and $\mathcal{L}_k$ with the property $\delta_k \in \mathcal{K}_k$ and $r_k - A \delta_k \perp \mathcal{L}_k$.

This is but a framework, in the same way that choosing some $M$ and $N$ to split $A$ into was a framework which described many methods as well as a general class of things you could do which might not be valuable at all. Consequently, our existing splitting methods can be described as projection methods. For instance, say for each step $k$ we set $\mathcal{K}_k = \mathcal{L}_k = \mathrm{span}(e_k)$ where $e_k \in \reals^n$ is the unit vector which is $e_k^i = 1$ for $i = k$ and $e_k^{i} = 0$ otherwise. In this case, our method appears as so \begin{align*}
\delta_k \in \mathrm{span}(e_k) &\implies \delta_k = c e_k \\
r_k - A \delta_k \perp \mathrm{span}(e_k) &\implies (r_k - c A e_k ) \cdot e_k = 0 \\
& \implies c = \frac{r_k \cdot e_k}{e_k \cdot A e_k}
\end{align*}
where we state the orthogonality condition as a dot product equality and rearrange to obtain $c$. Recalling $r_k = b - A x_k$, we can state each step as a standard recurrence relation \begin{align*}
x_{k+1} = x_k + \delta_k &= x_k + e_k \frac{(b - A x_k) \cdot e_k}{e_k \cdot A e_k}.
\end{align*}
Index notation is especially ideal for expressing this since each dot product by $e_k$ is equal to the $k$'th element of the other argument, so $e_k \cdot A e_k$ is $A_{kk}$ and $(b - Ax_k) \cdot e_k = b^k - \sum_j A_{kj} x_k^j$. However these only contribute to the coefficient $c$ seen here as the fraction; the vector $e_k$ which is scaled by this fraction is zero if the index we are inspecting is not itself $i = k$. So \begin{align*}
x_{k+1}^k &= x_k^k + \frac{b^k - \sum_j A_{kj} x_k^j}{A_{kk}} \\
 &= \frac{1}{A_{kk}} \left( A_{kk} x_k^k + b^k - \sum_{j=1}^n A_{kj} x^j_k \right).
\end{align*}
Consider now that since each step updates one element of $x_k$ at a time, all $x_k^j$ with $j > k$ will refer to the unaltered vector $x_0$. Additionally, each $x_k^j$ with $j < k$ will only have been altered once by this iteration scheme; in general on  any iteration step $x_{k+1}$, we are in the process of altering element $k$ and thus $x_k^k = x_0^k$. If we denote $y = x_n$ i.e. each element $y^k$ refers to the post-alteration $x^k_{k+1}$ with the full vector $y$ being the result after $n$ projection steps, then in fact this is \begin{align*}
x_{k+1}^k &= \frac{1}{A_{kk}} \left( A_{kk} x_0^k + b^k - \sum_{j<k}^n A_{kj} x^j_k - \sum_{j>k}^n A_{kj} x^j_0 \right) \\
y^k &= \frac{1}{A_{kk}} \left( A_{kk} x_0^k + b^k - \sum_{j<k}^n A_{kj} y^j - \sum_{j>k}^n A_{kj} x^j_0 \right)
\end{align*}
at which point it becomes apparent that the $n$ projection steps of this projection method which move us from $x_0$ to $y = x_n$ is in fact the same form as Gauss-Seidel iteration. For $\mathcal{K}_k = \mathcal{L}_k = \mathrm{span}(e_k)$, they produce the same algorithm.

But the value of the projection method is lost on us if all we do is choose subspaces which are of single dimension. In such a case, projection methods are no more valuable than splitting methods, perhaps even less so because of the difficulties in converting to a clean iteration.

To illustrate where projection methods offer us something novel, assume we have a $Ax = b$ problem with $A \in \reals^{n \times n}$ and $n = 2m$ i.e. is divisible by two, and set $\mathcal{K}_k = \mathcal{L}_k = \mathrm{span}(e_{2m-1}, e_{2m})$ so that each subspace $\mathcal{K}_k = \mathcal{L}_k$ has two dimensions spanned by consecutive orthonormal unit vectors. In this case, we are in some sense trying to approximate Gauss-Seidel but seeing if we can do it two steps at a time, and the projection method indicates an iteration that satisfies the following:
\begin{align*}
&\delta_k \in \mathrm{span}(e_{2k-1}, e_{2k}) \implies \delta_k = c_1 e_{2k-1} + c_2 e_{2k} \\
& r_k - A \delta_k \perp \mathrm{span}(e_{2k-1},e_{2k}) \implies \begin{bmatrix} (r_k - A \delta_k) \cdot e_{2k-1} \\ (r_k - A \delta_k) \cdot e_{2k} \end{bmatrix} = \begin{bmatrix} 0 \\ 0\end{bmatrix}\\
& \implies \begin{bmatrix} e_{2k-1} \cdot r_k \\ e_{2k} \cdot r_k \end{bmatrix} -
\begin{bmatrix}
	e_{2k - 1} \cdot A e_{2k-1} & e_{2k-1} \cdot A e_{2k} \\
	e_{2k} \cdot A e_{2k-1} & e_{2k} \cdot A e_{2k}
\end{bmatrix}
\begin{bmatrix} c_1 \\ c_2 \end{bmatrix} = \begin{bmatrix} 0 \\ 0 \end{bmatrix} \\
& \implies \begin{bmatrix} c_1 \\ c_2 \end{bmatrix} = \begin{bmatrix}
	e_{2k - 1} \cdot A e_{2k-1} & e_{2k-1} \cdot A e_{2k} \\
	e_{2k} \cdot A e_{2k-1} & e_{2k} \cdot A e_{2k}
\end{bmatrix}^\inv
\begin{bmatrix} e_{2k-1} \cdot r_k \\ e_{2k} \cdot r_k \end{bmatrix}
\end{align*}
This is exactly the same as before, where we converted the orthogonality condition into a dot product, but this time there are two dot product conditions which must be simultaneously obeyed, $(r_k - A \delta_k) \cdot e_{2k-1} = 0$ and $(r_k - A \delta_k) \cdot e_{2k} = 0$ and in addition to that, $\delta_k$ is now a linear combination of both $e_{2k-1}$ and $e_{2k}$. Here, we have expanded $\delta_k = c_1 e_{2k-1} + c_2 e_{2k}$ and then expressed the simultaneous dot products we require as a linear equation in $\reals^2$

Calculating the proper $c_1$ and $c_2$ such that $\delta_k$ satisfies $\delta_k \in \mathcal{K}_k$ and $r_k - A\delta_k \perp \mathcal{L}_k$ is now more difficult, but not so much more difficult since we have a linear relation. Yes, we require a matrix inversion, but the matrix inversion is only $2 \times 2$, where the typical formula \begin{gather*}
\begin{bmatrix} a & b \\ c & d\end{bmatrix} ^ \inv = \frac{1}{ad - bc} \begin{bmatrix} d & -b \\ -c & a\end{bmatrix} ^ \inv
\end{gather*}
will suffice. In this case, we can decide how large a matrix we are comfortable inverting by other means, whether that be $2\times 2$, $3 \times 3$, $p \times p$ with $p \ll n$, etc., and if for some reason we think we can do these calculations quicker, we can remove vast swathes of error from our residual all at once. Moreover, if we know something about where our error is coming from, we may be able to select our $\mathcal{K}$ and $\mathcal{L}$ to target and subtract large sources of error directly, something we couldn't do when we were bound to repeat only the same splitting method iteration.

So far, we have only examined cases where $\mathcal{K} = \mathcal{L}$ and an orthonormal basis of these subspaces are immediately available, principly as unit vectors $e_i$. If we get more creative with our choices of subspaces we will need to break the latter assumpion, but we will soon see that assuming some basis exists and aiming to make its calculation part of our iteration process later will be valuable. Assuming a basis $\{v_1, ..., v_p \}$ for $\mathcal{K}$ and $\{w_1, \dots, w_p\}$ for $\mathcal{L}$, then forming the matrices $V$ and $W$ whose columns are these bases (with shape $n \times p$), our technique appears as follows: \begin{enumerate}
\item Calculate $r_k = b - Ax_k$.
\item If we are to express our earlier condition in terms of these basis vectors, it would appear we need to solve \begin{gather*}
w_i \cdot (r_k - A\delta_k) = 0 \hspace{2em} 1 \le i \le p
\end{gather*}
however with $\delta_k = c_1 v_1 + ... + c_p v_p$ it is prudent to construct a coefficient vector $\vec c$ for which $\delta_k = V \vec c$ and to combine all $p$ equations using the matrix $W$ (transposed so that we get a dot-product on each column), and so our condition becomes
\begin{gather*}
W^T (r_k - AV \vec c) = \vec 0 \\
\implies \vec c = (W^T A V)^\inv W^T r_k
\end{gather*}
with $(W^T A V)^\inv$ a $p \times p$ matrix inversion.
\item Finally, recall that $\vec c$ was the coefficients for how much of each basis vector $v_i$ we wanted to put in $\delta_k$, so our update step is \begin{gather*}
x_{k+1} = x_k + V \vec c.
\end{gather*}
\end{enumerate}

This notion of considering our update vector as an assembly of other vectors in another basis will become very important later.

For now, we direct our attention towards two important classes of projection techniques. So far we have considered the case where $\mathcal{L} = \mathcal{K}$, and here we can show that if our $A$ is symmetric, pretty much any choice of subspace $\mathcal{K} = \mathcal{L}$ will improve our guess. These kinds of methods with $\mathcal L = \mathcal K$ are called \textbf{orthogonal projection methods} (otherwise, the projection method is called \textbf{oblique}).

\begin{label proposition}{iterpageOrthogonalProjectionBasicOptimality}
Let $A \in \reals^{n \times n}$ satisfy $A = A^T$ and $v^T A v > 0$ for all $v \neq 0$, and say for some $b \in \reals^n$ we are trying to solve $A x_* = b$ for $x_* \in \reals^n$. If we have an initial guess $x$ and select some subspace $\mathcal{K} = \mathcal{L}$ of $\reals^n$, any $y = x + \delta$ with \begin{gather*}
\delta \in \mathcal{K} = \mathcal{L} \\
b - Ay \perp \mathcal{L} = \mathcal{K}
\end{gather*}
and define an error function \begin{gather*}
E(z) = \sqrt{(z - x_*) \cdot A(z - x_*)}.
\end{gather*}
Then $E(y) = \min_{\delta \in \mathcal K} E(x + \delta)$.
\end{label proposition}

\begin{my proof}{m}
Since $A$ is symmetric positive definite, it defines an inner product $\langle u,v\rangle_A = u \cdot A v$ and thus $E(z)$ defines a norm $\lVert z \rVert_A = \sqrt{\langle z, z \rangle_A}$. Note also that \begin{align*}
E(z) &= \sqrt{(z - x_*) \cdot (Az - x_*)} \\
 &= \sqrt{(z - x_*) \cdot (Az - b)} \\
 &= \sqrt{(z - x_*) \cdot r_z}
\end{align*}
where $r_z$ is the associated residual $Az - b$.

With this norm and inner product in hand, now recall the Galerkin method. Unfortunately it has not been discussed on this website before, but it is a common method for selecting an error minimizing vector from a subspace. The outline of the method, in the language of our present problem, is that if you can control $\delta \in \mathcal K$ and aim to minimize $r \in \reals^n$ (a residual) then this is achieved at the point in $\mathcal{K}$ for which the error vector points orthogonally out of the plane of possible choices; the reasoning here is that if at any point the error vector is not orthogonal, there is a direction you can move in which would reduce your error.

Tangibly, the statement of the Galerkin method is that to find $\delta$ such that $r \cdot r$ is minimized (i.e. $\min_{\delta \in \mathcal{K}} (r \cdot r)$) we must select $\delta$ that satisfies \begin{gather*}
r_\delta \cdot v = 0 \hspace{2em} \forall v \in \mathcal{K}.
\end{gather*}
with $r_\delta = b - A(x + \delta)$. It is clear however that this condition is the same as $b - A(x + \delta) \perp \mathcal{K}$ as called for by the projection method, so by applying the projection method, we are also applying the Galerkin method.

Since we have a valid inner product $\langle \square , \square \rangle_A$, this too defines a Galerkin method such that if what we are trying to minimize is not $r_\delta = b - A(x + \delta)$ but instead $x + \delta - x_*$, which we know to be the same minimization problem, the method is\begin{gather*}
\delta \in \mathcal{K} \text{ minimizes } \langle x + \delta - x_* , x + \delta - x_* \rangle_A \\
\text{when } \langle x + \delta - x_* , v \rangle_A = 0 \hspace{1em} \text{for all } v \in \mathcal{K}
\end{gather*}
Since we know minimizing $x + \delta - x_*$ to be the same objective as minimizing $b - A(x + \delta)$, we can ignore the requirement, but notice that the objective can be written as \begin{align*}
\langle x + \delta - x_* , x + \delta - x_* \rangle_A &= (x + \delta - x_*) \cdot A (x + \delta - x_*) \\
&= \big( E(x + \delta) \big)^2 
\end{align*}
And since the square and the square root are monotonic functions, we know that in fact what we find is \begin{gather*}
E(y) = \min_{\delta \in \mathcal K} E(x + \delta)
\end{gather*}
where $\delta$ is the iteration vector satisfying the projection method requirements.
\end{my proof}

The other case that will be of interest, breaking from our tradition so far, is when $\mathcal L = A \mathcal{K}$ i.e. $\mathcal L$ is the vector subspace defined by the image of $A$ on the subspace $\mathcal K$. In fact it is the ability to satisfy this class condition that will make Krylov subspaces of such use to us, rather than any particular \squote{interesting shape} that one might imagine Krylov subspaces having.

\begin{proposition}
Let $A \in \reals^{n \times n}$ and say for some $b \in \reals^n$ we are trying to solve $A x_* = b$ with $x_* \in \reals^n$. If we have an initial guess $x$ and select some subspaces $ \mathcal{L} = A \mathcal{K}$ of $\reals^n$, any $y = x + \delta$ with\begin{gather*}
\delta \in \mathcal{K} \\
b - Ay \perp \mathcal{L} = A \mathcal{K}
\end{gather*}
satisfies \begin{gather*}
\lVert b - Ay \rVert = \min_{\delta \in \mathcal{K}} \lVert b - A(x + \delta) \rVert.
\end{gather*}
\end{proposition}

\begin{my proof}{m}
To show this, it will be helpful to have a lemma. Call $P$ an orthogonal projector when it has the property that \begin{gather*}
x = Px + (I - P)x
\end{gather*}
for all $x \in \reals^n$ and say it is an orthogonal projector \emph{onto} $\mathcal{K}$ when $Px \in \mathcal{K}$ for all $x \in \reals^n$. Orthogonal projectors allow us to separate $x$ into its $\mathcal{K}$ and non-$\mathcal K$ components, meaning \begin{align*}
x \cdot x &= (P x + (I - P)x) \cdot (Px + (I - P) x) \\
&= Px \cdot Px + (I - P) x \cdot (I - P) x + 2 Px \cdot (I - P) x \\ 
&= Px \cdot Px + (I - P) x \cdot (I - P) x + 2 Px \cdot x - 2 Px \cdot Px \\
&= Px \cdot Px + (I - P) x \cdot (I - P) x + (2 P - 2P )x \cdot Px \\
&= Px \cdot Px + (I - P) x \cdot (I - P) x 
\end{align*} by bilinearity. Written as norms, this is \begin{gather*}
\lVert x \rVert^2 = \lVert Px \rVert^2 + \lVert (I - P)x \rVert^2
\end{gather*} i.e. the squared magnitudes separate cleanly with no cross terms. First, it is valuable for us to know that \begin{gather*}
\min_{y \in \mathcal K}\lVert x - y \rVert = \lVert x - Px \rVert
\end{gather*}
for all $x \in \reals^n$. The reason for this is that a candidate $y \in \mathcal K$ has $Px - y \in \mathcal K$ and $x - Px \perp \mathcal K$, meaning for a similar property as above, \begin{gather*}
\lVert x - y \rVert^2 = \lVert x - Px + Px -  y \rVert^2 = \lVert x - Px\rVert^2 + \lVert Px - y \rVert^2
\end{gather*}
Since $\lVert Px - y \rVert^2 \ge 0$ as a norm, we conclude $\lVert x - Px\rVert^2$ is the lowest we can get $\lVert x - y \rVert^2$ by setting $y = Px$ so that $\lVert Px - y \rVert^2 = 0$.

With this lemma in hand, we immediately deduce a corollary that \begin{gather*}
\min_{y_* \in \mathcal K} \lVert x - y_* \rVert = \lVert x - y \rVert
\end{gather*}
if and only if both $y \in \mathcal K$ and $x - y \perp \mathcal K$, since this implies $y = Px$.

Now return to our problem. Restating our conditions in a different form, we say a satisfactory $\delta_*$ for the projection method is one which is \begin{gather*}
\delta_* \in \mathcal{K} \\
(b - Ax) - A\delta_* \perp \mathcal L = A \mathcal K
\end{gather*}
and we want to show that this is equivalent to \begin{gather*}
\min_{\delta \in \mathcal K} \lVert (b - Ax) - A \delta \rVert = \lVert (b - Ax) - A\delta_* \rVert.
\end{gather*}
But we may just as well say that we require $(A\delta) \in \mathcal L = A \mathcal K$ instead of $\delta \in \mathcal K$, in which case what we are trying to prove is \begin{gather*}
\min_{(A\delta) \in \mathcal L} \lVert (b - Ax) - (A \delta) \rVert = \lVert (b - Ax) - A\delta_* \rVert.
\end{gather*}
In this form, it is also clear that we can restate the requirements on $\delta_*$ as \begin{gather*}
A \delta_* \in \mathcal{L} \\
(b - Ax) - A\delta_* \perp \mathcal L
\end{gather*}
and thus our goal statement is proved exactly by the earlier mentioned corollary, with its $y$ as $A \delta$ and its $x$ as our $(b-Ax)$.
\end{my proof}

\subsection{Krylov Subspace Methods}

(This section will largely follow the narration of Saad. While he I found him often hard to follow, ultimately it became clear that his treatment of Krylov methods had the most apparent through-line.)

As alluded to in the final part of the previous section, there is something special about a projection method in which we select $\mathcal L = A \mathcal K$ which we aim to exploit. Moreover, at this stage we have not discussed how a projection method might calculate a basis for its relevant subspaces so that we could describe an update step in terms of a coefficient calculation $\vec c = (W^T A V)^\inv W^T r$. 

We aim to solve these together. The first relatively minor insight to make is that with $\mathcal L = A \mathcal K$, strictly speaking we only need one basis, since each vector $v_i \in \mathcal K$ which forms a basis in $\mathcal K$ will form a basis $A v_i \in \mathcal L$. We can exploit this even more so if we choose $\mathcal K$ to be \begin{gather*}
\mathcal{K}_m (A,r_0) = \mathrm{span}(r_0, Ar_0, \dots, A^{m-1} r_0)
\end{gather*}
using the residual of our initial guess $r_0 = b - Ax_0$ to start the sequence of basis vectors. Now, these vectors are not at all guarenteed to be orthogonal, and actually they are not even guarenteed to be linearly independent; we imagine that they are linearly independent since if we were to find some $c A^p r_0 = r_0$, this would mean we have actually just solved our $Ax = b$ problem early, and we would take \begin{align*}
c A^p (b - Ax_0) - (b - A x_0) &= 0 \\
b - A (- c A^{p-1}b + cA^p x_0 - x_0) &= 0 \\
x_* = - c A^{p-1}b + cA^p x_0 - x_0
\end{align*}
and so this failure of linear independence is in fact an ideal outcome so long as we include early terminations in our algorithms. If our linear dependence is more complicated $A^p r_0 = c_1 A^k r_0 + c_2 r_0$ for instance, the solution follows as an appropriate linear combination \begin{gather*}
x_* = -A^{p-1}b + A^p x_0 - A^{k-1}b - A^{k}x_0 - c_2 x_0
\end{gather*}
but it will always be some linear combination $A^0 r_0$ since if we have $A^p r_0 = c_1 A^k r_0 + c_2 A^{l}r_0$ then this implies we would have found $A^{p-l} r_0 = c_1 A^{k-l} r_0 + c_2 r_0$ earlier. If no such linear dependence is found, we are provided with the tools to proceed with the standard iterative approach we have been interested in so far.

I must pause at this stage and say that we've found some interesting leads, but nothing concretely useful at this point; that will come later. With Krylov subspaces I had a certain persistent impatience that I can only imagine is common when encountering them. Everyone seems to describe them as very important and there is perhaps an expectation that their importance can be made evident from a simple definition; unfortunately, the value of Krylov subspaces will come largely from the techniques that are to follow.

\subsubsection{Arnoldi's Method and Full Orthogonalization Method (FOM)}

The first order of business for us will be to explore what it would mean to find an orthogonal basis in $\mathcal{K}$. We already have a basis in a Krylov subspace by default, the standard $\mathcal{K}_m (A,r_0)$ assuming any existing linear dependence is exploited to terminate the problem by providing a solution early, but we have yet to make this basis orthogonal.

This can be achieved with the Gram-Schmidt algorithm; yes, it is clear that using this places us in another $O(n^3)$ problem, but as we will see, we either have no intention of actually running the algorithm we are about to describe until certain properties are shown that will reveal it to be $O(n)$.

The Arnoldi algorithm then proceed relatively similarly to Gram-Schmidt but with a seemingly minor, yet very important, asterisk: \begin{itemize}
\item In establishing our orthonormal basis $v_1, \dots, v_m \in \mathcal K_m (A, r_0)$, choose $v_1 = r_0 / \lVert r_0 \rVert$. Initialize the rest of the vectors $v_2, \dots, v_m$ as zeros such that the column vectors $v_1,\dots, v_m$ form the matrix $V$ which starts with only one non-zero column. Additionally, initialize a $v_{m+1} \in \reals^{n}$ as zeros.
\item Initialize a matrix $\tilde H$ which is $(m+1) \times m$ as zeros; we will soon refer to the top $m \times m$ part of the matrix as $H$ without the tilde. 
\item Initialize a sequence of buffer vectors $w_1, \dots, w_m \in \reals^n$ as zeroes.
\item For each $j = 1, \dots, m$, do the following: \begin{enumerate}
\item For each $i = 1, \dots, j$, set $\tilde H_{ij} = v_i \cdot A v_j$, thus filling out the first $j$ elements of the $j$'th column.
\item Set $w_j = A v_j - \sum_{i=1}^j \tilde H_{ij} v_i$ (this can be interpretted as $w_j = Av_j - V H e_j $ using the partially completed $V$ and $\tilde H$ and unit vector $e_j \in \reals^m$ which is $e_j^k = 1$ if $k=j$ and $0$ otherwise).
\item Set $\tilde H_{j+1,j} = \lVert w_j \rVert$, setting the element in $\tilde H$ just below the diagonal. If this $\lVert w_j \rVert = 0$, terminate the algorithm, as this is the linear dependence case mentioned earlier.
\item Set $v_{j+1} = w_j / H_{j+1, j} = w_j / \lVert w_j \rVert$.
\end{enumerate}
\end{itemize}
It is important to notice something about these $\tilde H$ and $H$ matrices we build. $H$, the top $m$ rows of $\tilde H$, is what is called a Hessenberg matrix, which is almost triangular except with non-zero elements just above or below the diagonal (below in our case). The reason for this in this instance is that what we have done here is actually slightly different from just Gram-Schmidt; if what we were doing matched Gram-Schmidt exactly, we would result in a triangular matrix of coefficients, but instead we have dropped the column which would have held the associated coefficient for $v_1$ (i.e. $\lVert r_0 \rVert$) and asserted our starting vector must be magnitude one. In place of this missing first column, we have constructed a basis vector $v_{j+1}$ which at the conclusion of our algorithm is $v_{m+1}$ and orthonormal to all our $v_1, \dots, v_m \in \mathcal{K}_m (A, r_0)$; in fact this is an orthonormal basis vector of $v_{m+1} \in \mathcal K_{m+1}(A,r_0)$, and if we included a column $[1,0,\dots,0]^T$ on the left of $\tilde H$, it would be the $(m+ 1) \times (m+1)$ diagonal matrix associated with the Gram-Schmidt algorithm.

By the conclusion of the algorithm, this means we have \begin{gather*}
A V = V H + w_{m} e_m^T
\end{gather*}
where $e_m$ is the $[0,\dots, 0, 1]^T \in \reals^m$ unit vector resulting in the outer product $w_m e_m^T$ an $n \times m$ matrix which is $w_m$ in the first column and zero otherwise. And this is owing exactly to the step \begin{gather*}
w_j = Av_j + V H e_j
\end{gather*}
since, at the final step, we can read this as \begin{gather*}
(w_m e_m^T) e_m = AV e_m  - V H e_m.
\end{gather*}
Using $\tilde H$ and with $v_{m+1} \in \mathcal{K}_{m+1} (A,r_0)$ as $w_m / \lVert w_m \rVert$ above used to write $\tilde V$, a $n \times (m+1) $ matrix whose columns are $v_1, \dots, v_m, v_{m+1}$, we can also write $AV = \tilde V \tilde H$, compressing the $w_m e_m^T$ into the $\tilde V \tilde H$ notation.

We have no intention of running this algorithm in its present form, since it involves a Gram-Schmidt orthogonalization of complexity $O(n^2)$. The takeaway from this for us is that, since $V$ is orthonormal and orthongal to $w_m$, the above property results in \begin{align*}
V^T A V &= V^T V H + V^T w_m e_m^T \\
&= I H + \mathbf{0} \\
&= H
\end{align*} 
where in particular this $\mathbf{0}$ is the zero matrix in $m \times m$. We conclude that $H = V^T A V$.

This is in fact sufficient to lead to \emph{a} method for solving $Ax=b$ problems, although it offers no efficiency until further modifications can be made. Observe that for an orthogonal projection method, that is, one where we use $\mathcal K = \mathcal L$, $V$ becomes a matrix forming a basis for both $\mathcal K$ and $\mathcal L$. Consequently our previous expression for the coefficient vector \begin{gather*}
\vec c = (W^T A V)^\inv W^T 
\end{gather*}
which solves the projection method problem with $x_{k+1} = x_k + V \vec c$ has $W = V$ and so \begin{align*}
\vec c &= (V^T A V)^\inv V^T r_k \\
&= H^\inv V^T r_k.
\end{align*}
Moreover, recalling that $V$ has columns which are the orthonormal vectors computed from Arnoldi's method, and in particular $v_1$ is the normalized $r_k$, $r_k$ is up-to-normalization the same as the first column of $V$. We read $V^T r_k$ as the dot product of the different columns of $V$ with $r_k$, where all columns are orthogonal except the first, so we get $V^T r_k = \lVert r_k \rVert e_1$ with $e_1 = [1,0,\dots,0]^T \in \reals^m$.

What we call the \textbf{Full Orthogonalization Method} is nothing but the Arnoldi method followed by the steps \begin{gather*}
\vec c = H^\inv (\lVert r_k \rVert e_1) \\
x_{k+1} = x_k + V \vec c
\end{gather*}
Where $H^\inv$ calls for an $m \times m$ matrix inverse. Curiously, we may write this as $\lVert r_k \rVert (H^\inv e_1) $ in which case it is clear that we are not meaning to calculate the entire matrix $H^\inv$, but only its first column; this could present a significant efficiency gain on a naive inversion algorithm, say, by using some kind of matrix adjugate algorithm perhaps, although this is not an avenue we will chase here.

Arnoldi's method, like all naive Gram-Schmidt implementations, is not generally robust to floating point rounding errors. There exists a modified Gram-Schmidt method consistent with the technique Arnoldi applies, but if for some reason you wish to implement this yourself, see algorithm 6.2 of Saad's aforementioned book.

\subsubsection{The Symmetric Lanczos Algorithm}

Now, if we were to find ourselves in a case where we had $A = A^T$ symmetric, this property we have just discovered where there exists a basis of $m$ orthonormal vectors in $\reals^n$ forming the matrix $V$ such that $H = V^T A V$ with $H$ a Hessenberg matrix means that \begin{gather*}
(V^T A V)^T = V^T A^T (V^T)^T = V^T A V
\end{gather*}
and thus $H = H^T$. This is a \textbf{huge} discovery, since we know by construction that each $j$'th column of $H$ is zero after the $(j+1)$'th element of the column; adding symmetry to this means that $H$ has shape \begin{gather*}
H = \begin{bmatrix} \alpha_1 & \beta_2 & & & \\ \beta_2 & \alpha_2 & \beta_3 & & \\ & \beta_3 & \ddots & \ddots & \\ & & \ddots & \alpha_{m-1} & \beta_m \\ & & & \beta_m & \alpha_m \end{bmatrix}
\end{gather*}
with $\alpha_1, \dots, \alpha_m \in \reals$ and $\beta_2, \dots, \beta_m \in \reals$ and empty spaces in the described matrix shape above filled with zeros.

In fact, this tells us that when we are running Arnoldi's algorithm, each $j$ iteration has to deal with at most three non-zero $v_i \cdot A v_j$, simplifying the runtime down to $O(n)$ vector operations. We call this the \textbf{Symmetric Lanczos's Algorithm}, and it runs as so: \begin{itemize}
\item In establishing our orthonormal basis $v_1, \dots, v_m \in \mathcal K_m (A, r_0)$, choose $v_1 = r_0 / \lVert r_0 \rVert$. Initialize the rest of the vectors $v_2, \dots, v_m$ as zeros, as well as a $v_0 = \vec 0 \in \reals^n$.
\item Initialize scalar values $\alpha_1, \dots, \alpha_m \in \reals$ and $\beta_1, \dots, \beta_m \in \reals$ as zeros, as well as buffer vectors $w_1, \dots, w_m \in \reals^n$. In particular, $\beta_1 = 0$ must be defined.
\item For each $j = 1, \dots, m$, do the following: \begin{enumerate}
\item Set $w_j = Av_j - \beta_j v_{j - 1}$. This can be considered the first part of the step that was $w_j = Av_j - \sum_{i=1}^j \tilde H_{ij} v_i$ in the Arnoldi algorithm, however we are still missing $\alpha_j$  in column $j$ so our $-\sum_{i=1}^j \tilde H_{ij} v_i$ cannot fully do its job yet. We will have to account for this missing term next.
\item Set $\alpha_j = w_j \cdot v_j$.
\item Update $w_j$ to  $w_j - \alpha_j v_j$. This completes the analogous $w_j = Av_j - \sum_{i=1}^j \tilde H_{ij} v_i$ step in the Arnoldi algorithm.
\item Set $\beta_{j+1} = \lVert w_j \rVert$. If $\lVert w_j \rVert = 0$ then we are in a linear dependence situation and must terminate the algorithm.
\item Set $v_{j+1} = w_{j}/ \beta_{j+1}$
\end{enumerate}
\end{itemize}

This performs the same process as the Arnoldi algorithm, which as we discussed earlier is no more than Gram-Schmidt, but by careful choice of which subspace we wanted to find and what kind of basis we wanted, we found that the Gram-Schmidt algorithm runs in this context in $O(n)$ time.

Once again, this algorithm works under the assumption of exact arithmetic, and in particular is known to fail rapidly under floating point rounding. Saad notes in his book:

\dquote{\emph{It is rather surprising that the above simple algorithm guarantees, at least in exact arithmetic, that the vectors $v_i$, $i = 1, 2, \dots ,$ are orthogonal. In reality, exact orthogonality of these vectors is only observed at the beginning of the process. At some point the $v_i$’s start losing their global orthogonality rapidly. There has been much research devoted to finding ways to either recover the orthogonality, or to at least diminish its effects by partial or selective orthogonalization}}

We will soon find remedies for this problem, but for now, we can notice that in exact arithmetic, we have yet another algorithm via applying FOM as in the previous subsection. Moreover, any technique that can give us efficiency gains on calculating the inverse of a tridiagonal matrix such as our symmetric $H$ will dramatically speed up such an FOM method.

\subsubsection{Direct Lanczos Algorithm}

A number of insights follow exactly the previous thread to enable the symmetric Lanczos FOM to turn into an efficient method. In the subsection after this one we will see about making it into an effective one.

First, we consider an $LU$-decomposition of our tridiagonal $H$. The techniques related to the inversion of banded matrices are discussed at length \hyperlink{https://doi.org/10.1016/j.cam.2012.07.018}{in Kılıç and Stanica 2013} (including an elegant proof that this decomposition works as described) and much more briefly in Saad, however for our purposes, all we need to know is that a tridiagonal matrix such as symmetric $H$ has an $LU$-decomposition in which $H$ is written\begin{gather*}
 \begin{bmatrix} \alpha_1 & \beta_2 & & & \\ \beta_2 & \alpha_2 & \beta_3 & & \\ & \beta_3 & \ddots & \ddots & \\ & & \ddots & \alpha_{m-1} & \beta_m \\ & & & \beta_m & \alpha_m \end{bmatrix} = 
\begin{bmatrix} 1 &  & & & \\ \lambda_2 & 1 &  & & \\ & \lambda_3 & \ddots & & \\ & & \ddots & 1 &  \\ & & & \lambda_m & 1 \end{bmatrix}
\begin{bmatrix} \eta_1 & \beta_2 & & & \\  & \eta_2 & \beta_3 & & \\ &  & \ddots & \ddots & \\ & & & \eta_{m-1} & \beta_m \\ & & &  & \eta_m \end{bmatrix}
\end{gather*}
Moreover, we have recurrence relations for these new quantities $\lambda_2,\dots, \lambda_m \in \reals$ and $\eta_1,\dots, \eta_m \in \reals$ \begin{gather*}
\lambda_j = \frac{\beta_j}{\eta_{j-1}} \\
\eta_j = \alpha_j - \lambda_j \beta_j
\end{gather*}
allowing for straight forward analytic calculation; this follows simply from gaussian elimination. Labelling these two matrices above $L$ and $U$ respectively, we take our earlier deduction that \begin{align*}
\vec c &= \lVert r_0 \rVert H^\inv e_1 \\
x_{k+1} &= x_k + V \vec c \\
&= x_k + \lVert r_0 \rVert V H^\inv e_1
\end{align*}
and rewrite it as \begin{gather*}
x_{k+1} = x_k + \lVert r_0 \rVert V U^\inv L^\inv e_1.
\end{gather*}
We now require a radical shift in our framing. So far we have privelaged $V$ as an orthonormal basis in $\mathcal K$ from which we build update vectors $\delta \in \mathcal K$ from a linear combination $\delta = V \vec c$. We now consider instead that the appropriate thing to calculate isn't the basis $v_1, \dots, v_m \in \reals^n$ exactly, but instead columns of a matrix we will call $P = V U^\inv$, together with a transformed coefficient vector $z = \lVert r_0 \rVert L^\inv \vec c$. At first it may seem that this only makes things more complicated, but here \hyperlink{https://doi.org/10.1016/j.cam.2012.07.018}{Kılıç and Stanica} can help us once again with the recurrence relation that defines the inverse of a bidiagonal matrix such as our $U$ or $L$. For $U$, this is \begin{gather*}
(U^\inv)_{ij} = \left\{ \begin{matrix} 0 & i > j \\ \frac1{\eta_j} \prod_{k=i}^{j-1} \frac{- \beta_{k+1}}{\eta_{k} } & i \le j \end{matrix} \right.
\end{gather*}
(Here I correct what I believe to be typos in the formula given in Kılıç and Stanica's paper. In particular what I write here is in place of their $v_j$ which should probably have been $u_j$ ((and indeed $v_j$ is not a meaningful symbol in their notation, hence why I am so sure)), and their product starts from $k=1$ whereas I believe it should start from $k=i$.)

A similar formula applies for $L^\inv$ since $L^T$ is an upper bidiagonal matrix and so $((L^T)^\inv)^T = L^\inv$ would mean this algorithm works there as well.
\begin{gather*}
(L^\inv)_{ij} = \left\{ \begin{matrix} 0 & i < j \\ 1 & i = j \\ \prod_{k=j}^{i-1} (- \lambda_{k+1}) & i > j \end{matrix} \right.
\end{gather*}

Alternatively, these can be expressed as a recurrence relation, which will allow us cut down on computations so long as we cache values as our hypothetical algorithm runs. \begin{gather*}
(U^\inv)_{ij} =  \left\{ \begin{matrix} 0 & i > j \\ \frac1{\eta_{j} } & i = j \\ \frac{- \beta_{i+1} }{\eta_{i} } (U^\inv)_{i+1,j} & i < j \end{matrix} \right. \\
(L^\inv)_{ij} =  \left\{ \begin{matrix} 0 & i < j \\ 1 & i = j \\ (- \lambda_{j+1}) (L^\inv)_{i,j+1} & i > j \end{matrix} \right.
\end{gather*}

If indeed we are to recontextualize $P$ and $z$ as the canonical objects of our algorithm rather than $V$ and $\vec c$, we are in a situation now where we need to develop intuitions about these objects. With our above recurrence relation, we can do just that.

Giving an example of our recurrence relation for $L^\inv$, in $m = 5$ it should look something like \begin{gather*}
L_5^{-1} = \begin{bmatrix} 1 & 0 & 0 & 0 & 0 \\ - \lambda_2 & 1 & 0 & 0 & 0 \\ \lambda_2 \lambda_3 & -\lambda_3 & 1 & 0 & 0 \\ -\lambda_2 \lambda_3 \lambda_4 & \lambda_3 \lambda_4 & -\lambda_4 & 1 & 0 \\ \lambda_2 \lambda_3 \lambda_4 \lambda_5 & - \lambda_3 \lambda_4 \lambda_5 & \lambda_4 \lambda_5 & - \lambda_5 & 1 \end{bmatrix}
\end{gather*}
This lets us do something really cool. Recalling that $z$ is defined $z = \lVert r_0 \rVert L^\inv e_1$, we know that this $z$ is simply this left hand column which is \begin{gather*}
z^j = \lVert r_0 \rVert \prod_{k=1}^{j-1} \lambda_{k+1}
\end{gather*}
or rather, we can think of a sequence of numbers $\zeta_j \in \reals$ with $\zeta_1 = \lVert r_0 \rVert$ and \begin{gather*}
\zeta_{j+1} = - \lambda_{j+1} \zeta_j
\end{gather*}
along with a sequence of vectors $z_j \in \reals^j$ which are similarly defined recursively as $z_1 = [\zeta_1]$ and \begin{gather*}
z_{j+1} = \begin{bmatrix} z_j \\ \zeta_{j+1} \end{bmatrix} \in \reals^{j+1}
\end{gather*}
via block matrix.

Where previously we had $\vec c$, a vector of coefficients we needed such that $V \vec c = \delta$ gave us the update vector we needed assuming we could calculate a single quite possibly expensive matrix inverse $H^\inv$, now, this transformed coefficient vector $z$ can actually be calculated \emph{iteratively}. That is to say, if we applied an orthogonal projection method with $\mathcal K = \mathcal L = \mathcal K_m (A, r_0)$ with a big $m$, previously this would have implied a big matrix inverse of $O(m^3)$; if we can make finding the $P$ matrix similarly easy, we have arrived at a $O(m)$ method for calculating the coefficient vector.

This discovery is of incredible significance to our framework of procedure. Where previously, the problem of a matrix inversion step meant that we might want to have a small $m$ and do many projection method steps, finding $x_{k+1} = x_k + \delta_k$ with $\delta_k \in \mathcal K$ and $b - A(x_k + \delta_k) \perp \mathcal L$, it no longer makes sense to separate these projection method steps. We are freed to pick a large $p$ and calculate in one algorithm \begin{gather*}
\tilde x = x_0 + \delta\\
\delta \in \mathcal K_p (A, r_0) \\
b - A \tilde x \perp \mathcal K_p (A,r_0)
\end{gather*}
now ditching subscripts of our iterative $x$ steps. It will of course still make sense to think of there being a sequence $x_m$, however these steps are no longer steps that occur due to repeated projection methods; from now on we operate within one projection algorithm until it terminates, and our $x_m$ will refer to how far along the projection method has progressed, equivalent to solving a projection method with $\mathcal K_{m}(A,r_0) = \mathcal K = \mathcal L$ for each $m \le p$.

In order to make this new system work as we are implying, we will need a way to iteratively come up with the columns of $P$. Since $P = V U^\inv$, we can look at an example $U^\inv$ in, say, $m=4$, with our above recurrence relation for the inverse: \begin{gather*}
U^\inv_4 = \begin{bmatrix} \eta_1^\inv & - \beta_2 (\eta_1 \eta_2)^\inv & \beta_2 \beta_3 (\eta_1 \eta_2 \eta_3)^\inv & - \beta_2 \beta_3 \beta_4 (\eta_1 \eta_2 \eta_3 \eta_4)^\inv \\
0 & \eta_2^\inv & -\beta_3 (\eta_2 \eta_3)^\inv & \beta_3 \beta_4 (\eta_2 \eta_3 \eta_4)^\inv \\
0 & 0 & \eta_3^\inv  & \beta_4 (\eta_3 \eta_4)^\inv \\
0 & 0 & 0 & \eta_4^\inv \end{bmatrix}
\end{gather*}
Paying attention to this, we notice that each $j+1$ column is merely the $j$'th column multiplied by $\beta_{j+1}\eta_{j+1}^\inv$ plus $\eta_{j+1}^\inv$ in the $(j+1)$'th row. Since we need to construct $P = V U^\inv$, and the first column of $U^\inv$ is zero in all but the first element, the first column of $P$ which we'll call $p_1$ is merely $\eta_1^\inv v_1$. Furthermore, chasing this recurrence relation, we'll see that each future column of $P$, $p_{j+1}$ is merely \begin{gather*}
p_{j+1} = \eta_{j+1}^\inv v_{j+1} - \beta_{j+1} p_j.
\end{gather*}

This is exactly what we wanted. Now, assuming exact arithmetic, we can run an iterative projection method at the same time that we run a basis finding algorithm such as Lanczos, in such a way that we can choose large $m$ and terminate at any time if our error becomes acceptable. Since the matrix $P$ adds a new column on each step, we'll denote $P_k \in \reals^{n \times k}$ to be the matrix on the $k$'th step with $k$ filled out columns; in this way, we have an update rule that looks like \begin{gather*}
x_k = x_0 + P_k z_k
\end{gather*}
remembering that $z_k \in \reals^k$. As an iterative rule, we only need to apply one of these columns at a time since each previous column will have been applied in $x_k$; we get \begin{gather*}
x_{k+1} = x_k + p_{k+1} \zeta_{k+1}.
\end{gather*}

The \textbf{Direct Lanczos} algorithm applies this as follows assuming $A$ is symmetric: \begin{itemize}
\item Choose an initial $x_0$ and compute an associated $r_0 = b - Ax_0$. Choose some maximum number of iterations $m \in \nats$ and initialize vectors $v_0,v_1, \dots, \in \mathcal K_m (A,r_0)$ with $v_0 = \vec 0$ and $v_1 = r_0 / \lVert r_0 \rVert $ as well as vectors $p_0,p_1, \dots, \in \reals^n$ with $p_0 = \vec 0$. We'll also need coefficient values $\alpha_1, \dots, \in \reals$, $\beta_1,\dots, \in \reals$ with $\beta_1 = 0$, $\eta_1,\dots, \in \reals$, $\lambda_1, \dots, \in \reals$ with $\lambda_1= 0$, and $\zeta_1,\dots, \in \reals$ with $\zeta_1 = \lVert r_0 \rVert$. Finally, we'll also use one buffer vector $w \in \reals^n$.
\item For each $k = 1, \dots, m$ do the following: \begin{enumerate}
	\item The first part of our loop will involve no more than following the Lanczos algorithm. Set $w = A v_k - \beta_k v_{k-1}$ just as before where it was the first part of an incomplete $w_k = A v_k - \sum_{i=1}^j \tilde H_{ij} v_i $ step as in the Arnoldi algorithm and described in the Lanczos algorithm.
	\item Set $\alpha_k = w \cdot v_k $. We will postpone the $w \to w - \alpha_k v_k$ update since it won't be necessary if we choose to terminate early.
	\item If $k > 1$ then set $\lambda_k = \beta_k \eta_{k-1}^\inv $ and $\zeta_k = -\lambda_k \zeta_{k-1}$. If $k=1$ then simply use $\lambda_1 = 0$ and $\zeta_1 = \lVert r_0 \rVert$ as described above.
	\item Set $\eta_k = \alpha_k - \lambda_k \beta_k$. Steps c. and d. together form the recurrence rule required to split $H = LU$ with $L$ and $U$ the bidiagonal matrices we defined earlier, as one derives from Gaussian elimination.
	\item Set $p_k = \eta_k^\inv (v_k - \beta_k p_{k-1})$. This is no more than the observation we made earlier that each $(j+1)$'th column of $U^\inv$ is but a product of its $j$'th column multiplied by $\beta_{j+1} \eta_{j+1}^\inv$ together with $\eta_{j+1}^\inv$ in its $(j+1)$'th row. This recurrence in $U^\inv$ means that $P = VU^\inv$ turns the $j$'th column into $p_j$ and an addition in a single row in $U^\inv$ into the $j$'th column of $V$ as seen above.
	\item Finally, set $x_k = x_{k-1} + \zeta_k p_k$. If $\lVert b - Ax_k \rVert$ is satisfactory, terminate.
	\item If $\lVert b - Ax_k \rVert$ is unsatisfactory, complete now the full update step mentioned earlier, updating $w$ to $w - \alpha_k v_k$.
	\item To prepare for the next step if indeed we have decided to continue, set $\beta_{k+1} = \lVert w \rVert$ and $v_{k+1} = w / \beta_{k+1}$. 
\end{enumerate}
\end{itemize}

Of course, a number of these steps may be skipped or moved around; for instance here we do not cache previous versions of $w$ since we no longer need to, and additionally it is up to the implementation circumstance as to how often one wants to check that $\lVert b - Ax_k \rVert$ is satisfactory. Similarly, it may also be unnecessary to keep previous $x_k$.

\subsubsection{The Conjugate Gradient Algorithm}

We now have an algorithm that, for symmetric $A \in \reals^{n \times n}$ and assuming exact arithmetic, will solve an $Ax=b$ problem in $O(n)$ vector operations. Although we have not emphasized it so far (since in practice this does not occur), under exact arithmetic and assuming $\mathcal K_n (A,r_0) = \reals^n$, this should solve for $x_*$ exactly on the $n$'th step. The trouble is that this algorithm depends on the Lanczos algorithm which is not stable in floating point precision; principly this is because we are running a straight forward naive Gram-Schmidt, obscured by a number of optimizations we have discovered.

Our task now will be to realize that features of this new algorithm are more robust than the particular methods we are using to calculate them.

\begin{proposition}
Let $r_k = b - A_k$ be the residual vectors produced by the Direct Lanczos algorithm above and thus inheriting its assumptions, along with the vectors $p_k$ and $v_k$ as produced by the algorithm. Then we have the following: \begin{enumerate}
\item Our residuals point exactly in the direction of our orthonormal vectors $v_k$, that is to say for some $\sigma_k \in \reals$, \begin{gather*}
r_k = \sigma_k v_{k+1}.
\end{gather*}
Consequently, $r_i \cdot r_j = 0$ when $i \neq j$.
\item The vectors $p_k$ are $A$-orthogonal as we described much earlier, i.e. they satisfy $p_i \cdot Ap_j = 0$ when $i \neq j$.
\end{enumerate}
\end{proposition}

\begin{my proof}{m}
The first claim follows from a much earlier fact about the Arnoldi method. Recall the equation $AV = VH + w_{m} e_m^T$, and that this was just $AV = VH + \beta_{m+1} v_{m+1} e_m^T$. Proceeding with terminology appropriate to the Arnoldi FOM algorithm (since indeed direct Lanczos is just that with some optimizations), observe that with $x_k = x_0 + V \vec c$ we have \begin{align*}
r_k = b - Ax_k &= b - A(x_0 + V \vec c) \\&= (b - Ax_0) - AV\vec c \\&= r_0 - VH\vec c - \beta_{m+1} v_{m+1}e_m^T \vec c
\end{align*}
We can take $e_m^T \vec c$ as a dot product, together with the factor $\beta_{m+1}$ to obtain $\sigma_m = - \beta_{m+1} e_m^T \vec c$. As for $r_0$ and $VH \vec c$, recall that $v_1 = r_0 / \lVert r_0 \rVert $ so $r_0 = \lVert r_0 \rVert v_1$; we also derived earlier when discussing the full orthogonalization method that $\vec c= H^\inv (\lVert r_0 \rVert e_1)$, which means that $H \vec c= \lVert r_0 \rVert e_1$. Substituting these in we get \begin{align*}
r_k &= \lVert r_0 \rVert v_1 - V(\lVert r_0 \rVert e_1) + \sigma_m v_{m+1} \\
&= \lVert r_0 \rVert v_1 - \lVert r_0 \rVert v_1 + \sigma_m v_{m+1} \\
&= \sigma_m v_{m+1}
\end{align*}
where we use $V e_1 = v_1$ since the $e_1 = [1,0,\dots,0]^T $ unit vector can extract the first column of the matrix. We are done with part a.

The second part is fundamentally a proof that $P^T A P$ is a diagonal matrix, since each element $(P^T A P)_{ij}$ correspond to the elementwise formula \begin{align*}
(P^T A P)_{ij} &= \sum_{k,l} (P^T)_{ik} A_{kl} P_{lj} \\
&= \sum_{k,l} P_{ki} A_{kl} P_{lj}
\end{align*}
and so since we sum across each row in $k$ and $l$, in effect what we are looking at in the $(i,j)$ position of $P^T A P$ is a dot product $p_i \cdot A p_j$.

Let us begin by using $P = VU^\inv$. Recalling $V^T A V = H$ and $H = LU$, we have \begin{align*}
P^T A P &= U^{-T} V^T A V U^\inv \\
&= U^{-T} H U^\inv \\
&= U^{-T} LU U^\inv \\
&= U^{-T} L.
\end{align*}
Now since $U$ is upper triangular, we must have $U^{-T}$ lower triangular. Moreover, triangular matrices form closed groups, so $U^{-T} L$ must also be lower triangular; however we also know that $P^T A P$ is symmetric via the symmetry of $A$, so $(P^T A P)^T = P^T A^T P = P^T A P$. If $U^{-T}L$ is lower triangular yet it is also symmetric, then it can only be non-zero on its diagonal. As above, this is the same as $p_i \cdot A p_j = 0$ for $i \neq j$. We are done with part b.
\end{my proof}

Knowing these properties, it now becomes appropriate to disassemble the direct Lanczos method in an effort to reproduce its artifacts without following the method itself. In particular, we know this must be possible in some manner since as long as we start with $p_0 = r_0$, each successive Krylov subspace $\mathcal K_2 (A , r_0) = \mathrm{span}(r_0,A r_0)$, then $ \mathrm{span}(r_0,Ar_0,A^2 r_0)$, will each specify a unique $A$-orthogonal direction at step $k+1$ not previously spanned by the $P$ basis in $k$; consequently, up to a scalar factor, we have no commitments to any way of calculating $p_{k+1}$ since it is well defined at each step. We know we cannot follow a naive Gram-Schmidt method or else we will produce an unstable method, but since in theory it is possible to derive some set of steps such as \begin{gather*}
x_{k+1} = x_k + \alpha_k p_k \\
p_{k+1} = \eta_k^\inv \sigma_{k}^{-1} r_k + \beta_k p_k 
\end{gather*}
we can proceed as follows as follows.

Since we are trying to dispense with Gram-Schmidt, we dismiss the $\alpha_k, \beta_k, \lambda_k$ symbols we had earlier which were artifacts of deriving a Gram-Schmidt $H$ matrix or its $LU$-decomposition; instead we use these symbols now to mean unknown coefficients which we need to find new ways to calculate, since e.g. what we now call $\alpha_k$ was previously $\zeta_k$.

By the above proposition, we have also learned that we can dismiss any explicit calculation of $v_k$ since they are a biproduct of the residuals of this method, i.e. via $\sigma_{k} v_{k+1} = r_k$ we derive $v_{k+1} = \sigma^\inv_{k} r_k$; moreover, the step we had previously in the direct Lanczos algorithm which was $p_k = \eta_k^\inv (v_k - \beta_k p_{k+1})$ takes $\eta_k^\inv$ and $v_{k+1} = \sigma^\inv_{k+1} r_k$ to give the first term of the formula above.

But since we are dismissing any attachment to $H$ or $V$, and we had only defined the vectors $p_k$ as the columns of $P = VU^\inv$, introducing a scalar multiple into each $p_k$ does not disrupt any of the properties we wanted our $p_k$ vectors to have. Consequently, we can dispense entirely with the $\eta_k^\inv \sigma_{k+1}^\inv$ factor, assuming $p_{k+1}$ now holds in it some appropriate scaling and expecting $\beta_k$ and $\alpha_k$ to be scaled appropriately as well.

This leaves us with only two steps and two quantities we must calculate: \begin{gather*}
x_{k+1} = x_k + \alpha_k p_k \\
p_{k+1} = r_k + \beta_k p_k 
\end{gather*}
In fact, we can deduce $\alpha_k$ and $\beta_k$ purely from the consequences of the above proposition, namely that $r_i \cdot r_j = 0$ and $p_i \cdot A p_j = 0$ when $i \neq j$. Substituting our recurrence relation for $x_k$ into the formula for residual, we have \begin{align*}
r_{k+1} = b - A(x_k + \alpha_k p_k) &= (b - Ax_k) - \alpha_k A p_k \\
&= r_k - \alpha_k A p_k
\end{align*}
Merely enforcing our condition that residuals are orthogonal, we can use $r_{k+1} \cdot r_k = 0$ to obtain \begin{align*}
0 &= r_k \cdot (r_k - \alpha_k A p_k) \\
&= r_k \cdot r_k - \alpha_k (r_k \cdot A p_k) 
\end{align*}
which can be rearranged to obtain \begin{gather*}
\alpha_k = \frac{r_k \cdot r_k}{r_k \cdot A p_k}.
\end{gather*}

A similar trick now applies to computing $p_k$ vectors since $p_{k+1}$ should be $A$-orthogonal to $p_k$. Writing $A p_k \cdot p_{k+1} = 0$ yields by the recurrence relation \begin{align*}
0 &= A p_k \cdot (r_{k+1} + \beta_k p_k) \\
&= A p_k \cdot r_{k+1} + \beta_k (A p_k \cdot p_k)
\end{align*}
which may be similarly rearranged to obtain \begin{gather*}
\beta_k = - \frac{r_{k+1} \cdot A p_k }{p_k \cdot A p_k}.
\end{gather*}
In this case, we'll use a reformulation of this since $r_{k+1} = r_k - \alpha_k A p_k $ so using $A p_k = \alpha_k^\inv (r_{k+1} - r_k)$ and $p_k = r_k + \beta_{k-1} p_{k-1} $ to express $\beta_k$ as \begin{align*}
\beta_k &= - \frac{\alpha_k^\inv r_{k+1} \cdot (r_{k+1} - r_k) }{ (r_k + \beta_{k-1} p_{k-1}) \cdot Ap_k} \\
&= - \frac{\alpha_k^\inv r_{k+1} \cdot (r_{k+1} - r_k) }{ r_k \cdot Ap_k} \hspace{2em} \text{(by $p_{k-1} \cdot A p_k = 0$)}\\
&= - \frac{\alpha_k^\inv r_{k+1} \cdot (r_{k+1} - r_k) }{ \alpha_k^\inv r_k \cdot (r_{k+1} - r_k)} \\
&= \frac{ r_{k+1} \cdot r_{k+1} }{ r_k \cdot  r_k} \hspace{2em} \text{(by $r_{k+1} \cdot r_k = 0$)}
\end{align*}

What we derive is a much simpler algorithm that unfortunately obscures the precise reason it works; as we have seen above, this is but a clever relabelling that avoids the pitfalls but ultimately does a similar job to a highly modified Gram-Schmidt, which due to previously discussed circumstances would be $O(n)$ complexity if not for compounding rounding error.

The final \textbf{Conjugate Gradient} algorithm is as follows: \begin{itemize}
\item Choose an initial $x_0$ and compute an associated $r_0 = b - Ax_0$. Initialize vectors $x_1,\dots, \in \reals^n$, $r_1,\dots, \in \reals^n$ and vectors $p_0,p_1,\dots, \in \reals^n$ with $p_0 = r_0$.
\item For $k = 1, \dots$ do the following: \begin{enumerate}
	\item Set $\alpha_k = \frac{r_k \cdot r_k}{p_k \cdot A p_k}$. (Since $r_k = p_k - \beta_{k-1} p_{k-1}$ and $p_k$ are $A$-orthogonal, this denominator is equivalent to the $r_k \cdot A p_k$ we called for above.)
	\item Set $x_{k+1} = x_k + \alpha_k p_k$
	\item Set $r_{k+1} = r_k - \alpha_k A p_k$
	\item Decide whether $\lVert r_{k+1} \rVert$ is tolerable and if so, terminate the algorithm.
	\item Set $\beta_k = \frac{r_{k+1} \cdot r_{k+1} }{r_k \cdot r_k}$
	\item Set $p_{k+1} = r_{k+1} + \beta_k p_k$
\end{enumerate}
\end{itemize}

It is important to keep in mind as we continue that this algorithm does something mathematically equivalent to the direct Lanczos algorithm, but computationally very different. The update vectors $p_k$ we deduce are exactly as in the direct Lanczos algorithm up to a scaling factor, but are simply calculated a different way that is less vulnerable to compounding rounding errors.

\subsection{Conjugate Gradient: Propeties and Modifications}

We cannot discuss the broader properties and situational improvements on conjugate gradient without first acknowledging that what we have described above is not universally agreed upon to be \emph{the} algorithm. The previous section follows from Saad's narrative, but Meurant provides one as well which I felt was, in his text, poorly motivated. It does however set the stage for how we can analyze the algorithm, so we'll need to pivot frameworks.

\subsubsection{Polynomial Splitting Techniques and Acceleration Methods}

While Saad derives the conjugate gradient method as a projection method, Meurant has a greater focus on splitting methods, which ultimately has its advantages.

Given a splitting method with $A = M - N$ and $M x_{k+1} = N x^k + b$, shortened with $B = M^\inv N$ and $c = M^\inv b$ to $x_{k+1} = Bx_k + c$, Meurant says that there may exist an alternative sequence $y_k$ with its own recurrence relation which can converge faster, built out of a linear combination of different $x_k$s at different stages of recurrence. Neverminding whether this works quite yet, such a sequence would be defined \begin{gather*}
y_k = \sum_{l=0}^k \alpha_{lk} x_l
\end{gather*}
where we require $\sum_{l=0}^k \alpha_{lk} = 1$ so that $y_k$ is always a weighted average of the previous $x_k$ and thus a sequence of $x_k = x_*$ will result in $y_k = x_*$.

If we examine the error of this $y_k$ over time, $\eta_k = x_* - y_k$, we find that \begin{align*}
\eta_k &= x_* - \sum_{l=0}^k \alpha_{lk} x_l \\
&= \sum_{l=0}^k \alpha_{lk} x_* - \sum_{l=0}^k \alpha_{lk} x_l \\
&= \sum_{l=0}^k \alpha_{lk} (x_* - x_l)
\end{align*}
Since $x_*$ satisfies $x_* = B x_* - c$, if we have $\varepsilon_k = x_* - x_k$ as our normal error, we see that the residual is \begin{gather*}
\varepsilon_k = x_* - x_k = Bx_* - c - (Bx_{k-1} - c) = B(x_* - x_{k-1}) = B \varepsilon_{k-1}
\end{gather*}
similar to back when we were doing proofs in the splitting section. Taking this as a recurrence relation, we obtain $\varepsilon_k = B^k \varepsilon_0$ and apply this to our weighted average to get \begin{gather*}
\eta_k = \sum_{l=0}^k \alpha_{lk} \varepsilon_l = \sum_{l=0}^k \alpha_{lk} B^l  \varepsilon_0.
\end{gather*}
What we see now appears to be a polynomial of order $k$ in $B$ with coefficients $\alpha_{lk}$ for the $l$'th order terms. We'll call this polynomial $P_k(t)$, i.e. \begin{gather*}
P_k(t) = \sum_{l=0}^k \alpha_{lk} t^l
\end{gather*}
and now write $\eta_k = P_k(B)\varepsilon_0$ as a matrix polynomial.

Now that we have deduced our $y_k$ error is a matrix polynomial with a set of coefficients, we may apply other methods that exist for polynomial minimization; in this case Meurant calls for the use of Chebyshev polynomials.

\begin{definition}[Chebyshev Polynomials]
We will denote $T_n(x)$ as \textbf{Chebyshev Polynomials}, that is, for integer $n\in \nats$ and $x \in [-1,1]$ they are polynomials that satisfy \begin{gather*}
T_n(x) = \cos (n \arccos x)
\end{gather*}
or alternatively for $t \in [a,b]$ they are\begin{gather*}
T_n(t) = \cos \left(n \arccos \left( \frac{a+b}{2} + \frac{b- a}{2} x\right) \right)
\end{gather*}
interpolating the original $|x| \le 1$ polynomial.
\end{definition}

Chebyshev polynomials are interesting because they are ostensibly polynomials which emulate in some sense a cosine bound to an interval, but with $n$ periods within the interval. Additionally, they can be defined with a recurrence relation as well as having very straight forward minimization properties.

We'll quote the following from Meurant's book verbatim:
\begin{theorem}
Let $T_n$ by a Chebyshev polynomial. Then, \begin{enumerate}
\item $T_0(x) = 1$, $T_1(x) = x$, \begin{gather*}
T_{n+1} (x) = 2x T_n(x) - T_{n-1}(x);
\end{gather*}
\item For $n \ge 1$, $T_n$ is an $n$-degree polynomial whose leading coefficient is $2^{n-1}$;
\item $T_n(-x) = (-1)^n T_n(x)$;
\item $T_n(x)$ has $n$ zeros in $[-1,1]$, namely \begin{gather*}
x_k = \cos \left( \frac\pi2 \frac{2k+1}{n} \right), \hspace{2em} k=0,1,\dots,n-1,
\end{gather*}
and $n+1$ extremas \begin{gather*}
x_k' = \cos \left( \frac{k\pi}{n} \right) \hspace{1em} \text{with} \hspace{1em} T_n(x_k') = (-1)^k, \hspace{1.7em} k=0,1,\dots,n
\end{gather*}
\item The Chebyshev polynomials are orthogonal with respect to the scalar product \begin{gather*}
\langle f, g \rangle = \int_{-1}^{+1}\frac{f(x)g(x)}{\sqrt{1-x^2}}dx
\end{gather*}
Moreover, \begin{gather*}
\langle T_i, T_j \rangle = \left\{ \begin{matrix} 0 & i \neq j \\ \frac{\pi}{2} & i = j \neq 0 \\ \pi & i = j = 0 \end{matrix} \right.
\end{gather*}
\item For all $n$-degree polynomials with leading coefficient $1$, $T_n/2^{n-1}$ has the smallest maximum norm, namely $1/2^{n-1}$
\end{enumerate}
\end{theorem}

For proof, Meurant cites \hyperlink{https://dl.acm.org/doi/abs/10.5555/574892}{Bjorck and Dahlquist's Numerical Methods (1990)}. While normally I would provide a summarization of the proof, I try not to get too involved with polynomial algebra at the moment; to proceed, we only need results a., d. and f. from the above.

That is, we are presently faced with a similar polynomial minimization problem, almost but not quite like that of part f. of the above theorem. We need to pick a polynomial $P_k$ such that $\lVert P_k(B) \rVert$ is minimized, in that we want parameters $\alpha_{lk}$ such that $\eta_k = P_k(B) \varepsilon_0$ will be minimized. Recall that $\lVert P_k(B) \rVert$ is the matrix norm, and so what we are trying to minimize is \begin{gather*}
\min_{P_k} \max_{v \neq \vec 0} \frac{\lVert P_k(B) v \rVert }{\lVert v \rVert}.
\end{gather*}

Supposing $B$ has eigenvalues within some range $[-\rho,\rho]$ with $\rho > 0$, then imagining we take $v$ to be the eigenvector associated with the largest eigenvalue $\lambda$, what we are looking at above is more like \begin{gather*}
\min_{P_k} \max_{\lambda \in [-\rho,\rho]} \lvert P_k(\lambda) \rvert
\end{gather*}

This is no longer a matrix polynomial minimization problem but a normal polynomial minimization problem, and so we can approach it with methods associated to the Chebyshev polynomials, owing to their ability to be minimal polynomials. Now our averaging property of the polynomial coefficients $\alpha_{lk}$ from earlier becomes important; we aren't just trying to find any minimal polynomial, we are trying to keep the property that $\sum_{l=0}^k \alpha_{lk} x_* = x_*$ so in effect we require $P_k(1) = 1$.

Using $P_k(1)=1$ and property f. above, our minimal polynomial should look something like $T_k(\lambda/\rho)/2^{n-1}$, sending inputs $\lambda \in [-\rho, \rho]$ to $T_k$'s natural domain $[-1,1]$, and normalize it by the $\lambda = 1$ case. This gives us \begin{gather*}
P_k(\lambda) = \frac{T_k(t/\rho)}{T_k(1/\rho)}.
\end{gather*}
Rewriting this as $T_k(\lambda/\rho) = T_k(1/\rho) P_k(\rho)$, we can use a. from the above theorem on $T_k(\lambda/\rho)$ to get \begin{gather*}
T_{k+1}(\lambda/\rho) = \frac{2\lambda}{\rho} T_k(\lambda/\rho) - T_{k-1}(\lambda/\rho) \\
\implies \\
T_{k+1}(1/\rho) P_{k+1}(\lambda) = \frac{2\lambda}{\rho} T_k(1/\rho) P_k(\lambda) - T_{k-1}(1/\rho) P_{k-1} (\lambda)
\end{gather*}

If indeed the above is a property that will follow from $P_k$ being a polynomial such that $\lVert P_k(B) \rVert$ is minimized, then without knowing the polynomial explicitly at this stage, we can replace $\lambda$ with $B$ (careful not to forget our $2\lambda/\rho$) and use $\eta_k = P_k(B) \varepsilon_0$ once again to obtain \begin{gather*}
T_{k+1}\left(\frac1\rho\right) P_{k+1}(B) \varepsilon_0 = \frac{2B}{\rho} T_k\left(\frac1\rho\right) P_k(B) \varepsilon_0 - T_{k-1}\left(\frac1\rho \right) P_{k-1} (B) \varepsilon_0 \\
\implies \\
T_{k+1}\left(\frac1\rho\right) \eta_{k+1} = \frac{2}{\rho} T_k\left(\frac1\rho\right) B \eta_k - T_{k-1}\left(\frac1\rho \right) \eta_{k-1}
\end{gather*}

From the properties of the necessary minimal polynomial, we have obtained a recurrence relation on $\eta_k$ and thus $y_k$ through $\eta_k = x_* - y_k$. \begin{gather*}
T_{k+1}\left(\frac1\rho\right) P_{k+1}(B) \varepsilon_0 = \frac{2B}{\rho} T_k\left(\frac1\rho\right) P_k(B) \varepsilon_0 - T_{k-1}\left(\frac1\rho \right) P_{k-1} (B) \varepsilon_0 \\
\implies \\
T_{k+1}\left(\frac1\rho\right) \eta_{k+1} = \frac{2}{\rho} T_k\left(\frac1\rho\right) B \eta_k - T_{k-1}\left(\frac1\rho \right) \eta_{k-1} \\
\implies \\
y_{k+1} - x_* = \left( \frac{2 T_k\left(\frac1\rho\right) }{\rho T_{k+1}\left(\frac1\rho\right) } \right)  B (y_k - x_*) - \left( \frac{T_{k-1} \left(\frac1\rho \right)}{T_{k+1}\left(\frac1\rho\right)} \right) (y_{k-1} - x_*)
\end{gather*}
We now need to do some term surgery on the above recurrence relation. The factor in the first term $-B x_*$ via $x_* = B x_* + c$ will become $-x_* + c$, and that $x_*$ along with its coefficient will participate in the following identity: with the original Chebyshev recurrence on $T_k(1/\rho)$ we can eliminate some of the coefficients on $x_*$ \begin{gather*}
T_{k+1} (1/\rho) = \frac{2}{\rho} T_k(1/\rho) - T_{k-1}(1/\rho) \\ \implies \\ - x_* = \frac{2 T_k (1/\rho)}{\rho T_{k+1} (1/\rho)} (-x_*) - \frac{T_{k-1}(1/\rho)}{T_{k+1}(1/\rho)} (-x_*)
\end{gather*}

These two modifications together allow us to eliminate $x_*$ and get something we might actually be able to calculate \begin{gather*}
y_{k+1} = \left( \frac{2 T_k\left(\frac1\rho\right) }{\rho T_{k+1}\left(\frac1\rho\right) } \right) (B y_k + c) - \left( \frac{T_{k-1} \left(\frac1\rho \right)}{T_{k+1}\left(\frac1\rho\right)} \right) y_{k-1}
\end{gather*}
given that Chebyshev polynomials are known and $\rho$ can be taken to be the magnitude of the largest eigenvalue of $B$, without knowing anything about the $x_k$ sequence. A further simplification is made on the right hand side by repeating the trick we used on for eliminating $x_*$ via the Chebyshev recurrence rule on $T_k(1/\rho)$ but in $y_{k-1}$. \begin{align*}
y_{k+1} &= \left( \frac{2 T_k\left(\frac1\rho\right) }{\rho T_{k+1}\left(\frac1\rho\right) } \right) (B y_k + c - y_{k-1} + y_{k-1}) - \left( \frac{T_{k-1} \left(\frac1\rho \right)}{T_{k+1}\left(\frac1\rho\right)} \right) y_{k-1} \\
&= \left( \frac{2 T_k\left(\frac1\rho\right) }{\rho T_{k+1}\left(\frac1\rho\right) } \right) (B y_k  + c - y_{k-1}) + \left( \frac{2 T_k\left(\frac1\rho\right) }{\rho T_{k+1}\left(\frac1\rho\right) } - \frac{T_{k-1} \left(\frac1\rho \right)}{T_{k+1}\left(\frac1\rho\right)} \right) y_{k-1} \\
&= \left( \frac{2 T_k\left(\frac1\rho\right) }{\rho T_{k+1}\left(\frac1\rho\right) } \right) (B y_k + c - y_{k-1}) + y_{k-1}
\end{align*}

If now, we can relabel that nasty Chebyshev polynomial coefficient on the first term to $\omega_{k+1}$ and find a recurrence relation on it, we will have a complete technique available to us. Setting \begin{gather*}
\omega_{k+1} = \frac{2 T_k\left(\frac1\rho\right) }{\rho T_{k+1}\left(\frac1\rho\right) } \hspace{2em} \left( \text{and thus } \frac{\rho}{2} \omega_k =\frac{T_{k-1}\left(\frac1\rho\right) }{T_{k}\left(\frac1\rho\right) } \right)
\end{gather*}
we can apply the recurrence relation on $T_{k+1}$ and derive \begin{align*}
\omega_{k+1} &= \frac{2 T_k\left(\frac1\rho\right) }{2 T_{k}\left(\frac1\rho\right) - \rho T_{k-1} \left(\frac1\rho\right)} \\
&= \frac{1 }{1 - \frac{\rho T_{k-1} (1/\rho)}{2 T_{k}(1/\rho)}} \\
&= \frac{1}{1 - \frac{\rho^2}{4}\omega_k}
\end{align*}

which gives us our recurrence in $\omega_k$. Referring back to part a. of our theorem above laying out the properties of Chebyshev polynomials, by $T_0(x) = 1$ and $T_1(x) = x$ we have have $\omega_1 = 1$ via our formula for $\omega_{k+1}$ above.

This provides what Meurant informally called an \textbf{Acceleration technique} and formally called \textbf{Chebyshev semi-iteration}. It calls for entirely dismissing the sequence $x_k$ which we would have iterated $x_{k+1} = Bx_k +c $ and instead iterating a weighted combination of $x_k$ which we call $y_k$, obeying a similar convergence to $x_*$. The algorithm follows, although may not match exactly with literature: \begin{itemize}
\item Initialize $y_0,y_1,\dots, \in \reals^n$ with $y_0 = \vec 0 $ and $y_1$ chosen as an initial guess, as well as $z_1,\dots, \in \reals^n$ and $\omega_1, \dots, \in \reals$ with $\omega_1 = 1$.
\item For each $k = 1,2,3, \dots,$, do the following: \begin{enumerate}
\item Set $\omega_k = \frac{1}{1 - \rho^2 \omega_k / 4}$
\item Set $z_k = M^\inv (b - A y_k)$ in place of calculating $By_k + c$. Recall this is because $A = M - N$, $B = M^\inv N$ and $c = M^\inv b$, so in fact this is $z_k = c - (I - B)y_k$ and thus $z_k + y_k  = B y_k + c$.
\item With $z_k$, set \begin{gather*}
y_{k+1} = \omega_{k+1} (z_k + y_k - y_{k-1}) + y_{k-1}.
\end{gather*}
\end{enumerate}
\end{itemize}

\subsubsection{Preconditioned Conjugate Gradient From Polynomial Methods}

Recall earlier that we derived the Conjugate gradient method not by rearranging steps of the direct Lanczos method but by assuming the $A$-orthogonal basis with the properties the direct Lanczos method guarenteed were held, and nothing else. We will do something similar here, but using a different method as our base.

We begin with what is called the \textbf{Richardson method}, which at first may be thought of as a splitting method $A = M - N$ with $M = \alpha^\inv I$ and $N = \alpha^\inv I - A$. As all splitting methods, we may express it as the iteration $x_{k+1} = M^\inv N x_k + M^\inv b$ iteration \begin{align*}
x_{k+1} &= \alpha I (\alpha^\inv I - A) x_k + \alpha b \\
&= x_k - \alpha (b - Ax_k)
\end{align*}

Additionally, if it aids the convergence of the method, it is also normal to pick a different $M= \alpha^\inv \bar M $ so long as $\bar M$ is easily invertible or $\bar M x = y$ problems are easily solved (the same things we discussed of $M$ in splitting methods). For much of this section we will assume this works because $\bar M$ is invertible, however tricks such as what was used in Gauss-Seidel similarly solve this problem, as we'll describe later. Setting $N = M - A$, we get \begin{gather*}
\bar M x_{k+1} = \bar M x_k + \alpha (b - Ax_k)
\end{gather*}
in which case we say that the Richardson method is \textbf{preconditioned} with preconditioner $\bar M$. If we write this in our standardized iteration form $x_{k+1} = Bx_k + c$ owing from $x_{k+1} = M^\inv N x_k + M^\inv$ we see \begin{align*}
x_{k+1} &= \alpha \bar M^\inv (\alpha^\inv \bar M - A) x_k + \alpha \bar M^\inv b \\
&= (I - \alpha \bar M ^\inv A) x_k + \alpha \bar M ^\inv b.
\end{align*}
This $x_{k+1} = Bx_k + c$ may be plugged directly into the acceleration technique described above, in which case we obtain  \begin{gather*}
y_{k+1} = \omega_{k+1} (\bar M^\inv (I - \alpha A) y_k + \alpha \bar M^\inv b - y_{k-1}) + y_{k-1}.
\end{gather*}

In the above Chebyshev semi-iteration algorithm, we transformed $By_k + c$ into $z_k + y_k$ by setting $z_k = M^\inv (b - Ay_k)$ which was equivalent to $z_k = c - (I - B)y_k$; a similar transformation will be possible here if we pay close attention to $\alpha$. Using $B$ and $c$ as above but this time setting $z_k = \bar M^\inv (b - A)y_k$, we observe by adding $\alpha^\inv y_k$ to both sides that \begin{align*}
z_k + \alpha^\inv y_k &= \bar M^\inv (b - Ay_k) + \alpha^\inv y_k \\
&=  \bar M^\inv (b - Ay_k + \alpha^\inv \bar M y_k) \\
&=  \bar M^\inv (b + (\alpha^\inv \bar M - A) y_k ) \\
&=  \bar M^\inv b + \alpha^\inv ( I - \alpha M^\inv A )y_k \\
&= \alpha^\inv c + \alpha^\inv B y_k
\end{align*}
showing us that this definition leads to $\alpha z_k + y_k = By_k + c$. This can go right back into our acceleration method forming the following iteration.\begin{gather*}
z_k = \bar M^\inv (b - Ay_k) \\
y_{k+1} = \omega_{k+1} (\alpha z_k + y_k - y_{k-1}) + y_{k-1}
\end{gather*}

This is the method we will use as our base to rederive conjugate gradient as a polynomial technique, but in order to do so, we'll also need to vary $\alpha$ on each step. Since $y_k$ was the thing we needed to converge anyway, we'll relabel that back to $x_k$ and give our base method as so: \begin{gather*}
z_k = \bar M^\inv (b - Ax_k) = \bar M^\inv r_k\\
x_{k+1} = \omega_{k+1} (\alpha_k z_k + x_k - x_{k-1}) + x_{k-1} \\
\omega_{k+1} = ? \\
\alpha_k = ?
\end{gather*}

We should note two things before we continue.

Staying in line with the recurring theme, there is at this stage no guarentee that this is indeed a method from which we can rederive conjugate gradient. For that reason exactly, we derived conjugate gradient earlier as a series of optimizations on Gram-Schmidt, culminating in the direct Lanczos method, where we ultimately said there is a better way to calculate the $p_k$ vectors and $\alpha_k$ parameters.

What we have in place of a guarentee that conjugate gradient is possible here is the mere observation that this method already roughly looks like conjugate gradient. Recall our recurrence formula assuming successful calculation of $\alpha_k$ and $\beta_k$ were \begin{gather*}
x_{k+1} = x_k + \alpha_k^P p_k \\
p_{k+1} = r_{k+1} + \beta_k^P p_k.
\end{gather*}
marked $P$ since these coefficients are from the projection method conjugate gradient. If we set $\bar M = I$ and rearrange our iteration in $x_k$ to get $x_{k+1} - x_k = \alpha^P_k p_k$, we see immediately that our acceleration based method replaces $z_k$ with $r_k$ and looks like \begin{gather*}
x_{k+1} = \omega_{k+1} (\alpha_k r_k + \alpha_{k-1}^P p_{k-1} ) + x_{k-1}.
\end{gather*}
We see clearly that the bracketted terms look a lot like the update step for $p_k$, but we also have a $x_{k-1}$ to work with. If we could set $\omega_{k+1}$ such that \begin{align*}
\omega_{k+1} (\alpha_k r_k + \alpha_{k-1}^P p_{k-1}) &= \alpha_{k-1}^P p_{k-1} + \alpha_k^P p_k \\
&= \alpha_{k-1}^P p_{k-1} + \alpha_k^P (r_k + \beta_k^P p_{k-1} ) \\
&= \alpha_{k}^P r_k + (\alpha_{k-1}^P + \alpha_k^P \beta_k^P )p_{k-1}
\end{align*}
then we would have the standard projection conjugate gradient method via \begin{align*}
x_{k+1} &= (\alpha_{k-1}^P p_{k-1} + \alpha_k^P p_k) + x_{k-1} \\
&= x_k + \alpha_k^P p_k = x_{k+1}
\end{align*} 
exactly as it was before.

So indeed we actually have good reason to think we can rederive a different version of the same basic method here, and we'll recall that the way we derived the method the first time was by enforcing only the assumptions that $r_i \cdot r_j = 0$ for all $i \neq j$ and $p_i \cdot A p_j$ for all $i \neq j$.

Where our derivation will differ this time is that this time, we no longer explicitly deal explicitly in $p_k$ vectors, and we found a nice place to put a \emph{preconditioner} which will generalize the method a good bit. We had originally defined $M = \alpha^\inv \bar M$, but since our $\alpha$ is now varying on each step, we'll drop the pretense and $\bar M$ will be just called $M$ from now on. This means the first thing we'll need to ensure for our method is some kind of $z_k$ orthogonality, and in particular, we will go for $M$-orthogonality, since when $M = I$ this was the same thing as $r_i \cdot r_j = 0$ for all $i \neq j$.

Enforcing this property immediately leads to a formula for $\alpha_k$, since we can require $z_k \cdot M z_{k+1} = 0$ ; showing this however is a lot easier if we first notice something basic about our iteration method. If we take our formula for $x_{k+1}$ and substitute it into $b - Ax_{k+1}$ we get \begin{align*}
r_{k+1} &= b - Ax_{k+1} \\
&= b - A\omega_{k+1} (\alpha_k z_k + x_k - x_{k-1}) - Ax_{k-1} \\
&= - \omega_{k+1}(\alpha_k A z_k + Ax_k - Ax_{k-1}) + r_{k-1} \\
&=  - \omega_{k+1}(\alpha_k A z_k - b + Ax_k + b - Ax_{k-1}) + r_{k-1} \\
&= - \omega_{k+1}(\alpha_k A z_k - r_k + r_{k-1}) + r_{k-1} \\
Mz_{k+1} &= - \omega_{k+1}(\alpha_k A z_k - Mz_k + Mz_{k-1}) + Mz_{k-1}
\end{align*}
which we can substitute into the $z_k \cdot M z_{k+1} = 0$ requirement to get \begin{align*}
0 = z_k \cdot M z_{k+1} &= z_k \cdot Mz_{k-1} - \omega_{k+1} z_k \cdot (\alpha_k A z_k - Mz_k + Mz_{k-1}) \\
&= (1 - \omega_{k+1})(z_k \cdot Mz_{k-1}) - \omega_{k+1} (\alpha_k z_k \cdot A z_k - z_k \cdot Mz_k)
\end{align*}
Enforcing $z_{i} \cdot M z_j = 0$ for all $i \neq j$ again, this time on $z_k \cdot M z_{k-1} = 0$, we can eliminate the first term on the right hand side, deriving \begin{gather*}
\alpha_k = \frac{z_k \cdot M z_k }{z_k \cdot A z_k}
\end{gather*}
since the two terms in the brackets must sum to zero.

We can now do a similar trick for deriving $\omega_k$ by starting from $z_{k-1} \cdot M z_{k+1}$. We can reuse almost all of the equations above \begin{align*}
0 = z_{k-1} \cdot M z_{k+1} &= z_{k-1} \cdot Mz_{k-1} - \omega_{k+1} z_{k-1} \cdot (\alpha_k A z_k - Mz_k + Mz_{k-1}) \\
&= (1 - \omega_{k+1})(z_{k-1} \cdot Mz_{k-1}) - \omega_{k+1} (\alpha_k z_{k-1} \cdot A z_k - z_{k-1} \cdot Mz_k)
\end{align*}
and eliminate the $z_{k-1} \cdot M z_k$ term by $M$-orthogonality of our $z_k$ vectors. Our $\alpha_k z_{k-1} \cdot A z_k$ term remains however, and we continue with \begin{align*}
& 0 = z_{k-1} \cdot Mz_{k-1} + \omega_{k+1}(z_{k-1} \cdot Mz_{k-1} + \alpha_k z_{k-1} \cdot A z_k) \\
\implies &\omega_{k+1} = \frac{z_{k-1} \cdot M z_{k-1}}{ \alpha_k (z_{k-1} \cdot A z_k) + z_{k-1} \cdot M z_{k-1} } \\
\implies & \omega_{k+1} = \frac{1}{1 + \alpha_k (z_{k-1} \cdot A z_k) / (z_{k-1} \cdot M z_{k-1})}
\end{align*}
Recall that the assumptions that made the direct Lanczos algorithm possible were that $A$ was symmetric, and if we want to imagine we are applying such a method to a preconditioned system $M^\inv A x = M^\inv b$ then we also need that $M^\inv$ is symmetric and thus $M$ itself. This implies that the splitting that leads to the acceleration technique we are turning into the conjugate gradient method has $A = M - N$ with $N = M - A$ also symmetric, and Meurant tells us this also leads to an alternative (seemingly preferential) formula for $\omega_k$.

Using our earlier formula for $r_{k+1} = Mz_{k+1}$ let us write \begin{align*}
Mz_{k} &= - \omega_{k}(\alpha_{k-1} A z_{k-1} - Mz_{k-1} + Mz_{k-2}) + Mz_{k-2} \\
 &= - \omega_{k}(\alpha_{k-1} (M - N) z_{k-1} - Mz_{k-1} + Mz_{k-2}) + Mz_{k-2} \\
\end{align*}
Now, if we dot product in $z_k$, the left hand side will be spared but most terms on the right hand side will go to zero by $z_k$'s $M$-orthogonality, including the $M$ component of our $A z_{k-1}$. What we get is 
\begin{gather*}
z_k \cdot Mz_{k} = \omega_{k}\alpha_{k-1} (z_k \cdot N z_{k-1})
\end{gather*}
which by symmetry, lets us swap $z_k \cdot N z_{k-1}$ for $z_{k-1} \cdot N z_{k}$ and gives us an expression \begin{gather*}
z_{k-1} \cdot A z_{k} = z_{k-1} \cdot N z_{k} = \frac{z_{k} \cdot M z_{k}}{\omega_k \alpha_{k-1}}
\end{gather*}
and substituting this into $\omega_{k+1}$, \begin{gather*}
\omega_{k+1} = \left( 1 + \frac{\alpha_k (z_k \cdot M z_k )}{\omega_k \alpha_{k-1}(z_{k-1} \cdot M z_{k-1})}\right)^\inv
\end{gather*}
implying one less matrix multiplication so long as we are caching our $\omega_k$'s and $\alpha_k$'s.

In total, we collect these steps into Meurant's \textbf{Preconditioned Conjugate Gradient} method. It requires $A$ to be symmetric and a choice of symmetric preconditioner $M$ for which an inverse $M^\inv$ is known or easily computed. \begin{itemize}
\item Choose an initial guess $x_0$ (with $x_{-1}$ set to $x_0$ although original authors \hyperlink{http://i.stanford.edu/pub/cstr/reports/cs/tr/76/533/CS-TR-76-533.pdf}{Concus, Golub and O'Leary 1976} suggest it may be set arbitrarily) and initialize vectors $x_0, \dots, \in \reals^n$, $r_0, r_1, \dots \in \reals^n$, and $z_0, z_1, \dots, \in \reals^n$ with $x_0 = 0$. Initialize coefficients $\alpha_0, \dots \in \reals$ as well as $\omega_1, \dots \in \reals$ with $\omega_1 = 1$.
\item For each $k = 0, \dots,$ do the following: \begin{enumerate}
\item Set $r_k = b - Ax_k$ and $z_k = M^\inv r_k$ (more broadly, solve the $M z_k = r_k$ system), or as desired, use formula \begin{gather*}
r_{k} = r_{k-2} - \omega_{k}(\alpha_{k-1} A z_{k-1} - r_{k-1} + r_{k-2}) \\
z_{k} = z_{k-2} - \omega_{k}(\alpha_{k-1} M^\inv A z_{k-1} - z_{k-1} + z_{k-2})
\end{gather*}
or calculate this after step c. so it is stated in $r_{k+1}$ or $z_{k+1}$ with cached $A z_k$. 
\item Set \begin{gather*}
\alpha_k = \frac{z_k \cdot r_k}{z_k \cdot A z_k} \hspace{2em} \left(= \frac{z_k \cdot Mz_k}{z_k \cdot A z_k}  \right)
\end{gather*}
\item Set \begin{gather*}
\omega_{k+1} = \left( 1 + \frac{\alpha_k (z_k \cdot r_k )}{\omega_k \alpha_{k-1}(z_{k-1} \cdot r_{k-1})}\right)^\inv
\end{gather*}
\item Set \begin{gather*}
x_{k+1} = x_{k-1} + \omega_{k+1} (\alpha_k z_k + x_k - x_{k-1})
\end{gather*}
\end{enumerate}
\end{itemize}
There are options here for whether you compute $z_k = M^\inv r_k$ on each step or set $B = M^\inv A$ and $c = M^\inv$ and instead calculate $z_k = c - Bx_k$ directly, then computing $Mz_k$ on each step.

Taking our lessons here in preconditioning back to Saad's conjugate gradient method, we can reformulate it in a way that uses $p_k$ vectors as well as a preconditioner $M$. Meurant informally calls this \dquote{second form conjugate gradient}. \begin{itemize}
\item Choose an initial $x_0$ and compute an associated $r_0 = b - Ax_0$. Initialize vectors $x_1,\dots, \in \reals^n$, $r_1,\dots, \in \reals^n$, $z_1,\dots, \in \reals^n$ and vectors $p_0,p_1,\dots, \in \reals^n$ with $p_0 = r_0$.
\item For $k = 0, \dots$ do the following: \begin{enumerate}
	\item Set $\alpha_k = \frac{z_k \cdot r_k}{p_k \cdot A p_k} \hspace{1em} \left( = \frac{z_k \cdot Mz_k}{p_k \cdot A p_k} \right)$.
	\item Set $x_{k+1} = x_k + \alpha_k p_k$
	\item Set $r_{k+1} = r_k - \alpha_k A p_k$
	\item Decide whether $\lVert r_{k+1} \rVert$ is tolerable and if so, terminate the algorithm.
	\item Set $\beta_{k+1} = \frac{z_{k+1} \cdot r_{k+1} }{z_k \cdot r_k } \hspace{1em} \left( = \frac{z_{k+1} \cdot Mz_{k+1} }{z_k \cdot Mz_k } \right)$
	\item Set $p_{k+1} = r_{k+1} + \beta_{k+1} p_k$
	\item Set $z_{k+1} = M^\inv r_{k+1}$
\end{enumerate}
\end{itemize}

At this stage, we have a number of options for different ways we can run this algorithm, picking which quantities we compute and when to run effectively the same method depending on our constraints.

\subsubsection{Optimality of Preconditioned Conjugate Gradient}

Much earlier we discussed in proposition \ref{pro:iterpageOrthogonalProjectionBasicOptimality} that projection techniques minimized a notion of error where $A$ defines a notion of distance from the true solution \begin{gather*}
E(x) = \sqrt{(x_* - x)^T A (x_* - x)}.
\end{gather*}
We found that orthogonal projection techniques maximally optimize this error within the subspace of allowable changes $\delta = x_{k+1} - x_k$, i.e. \begin{gather*}
E(x_{k+1}) = \min_{\delta \in \mathcal K} E(x_k + \delta). 
\end{gather*}

What we will see now is that preconditioned conjugate gradient in the form provided by Meurant (noting of course that this must be a mathematically equivalent process to orthogonal projection conjugate gradient and by extension direct Lanczos) is also maximally optimized in the sense of an accelerating averaging just like one that defined Chebyshev semi-iteration.

To show this, we'll need to specify a bit of what we mean. First, with $B = M^\inv A$ and our previous formula for $x_{k+1}$ iteration in terms of $z_{k+1}$, \begin{align*}
Mz_{k+1} &= - \omega_{k+1}(\alpha_k A z_k - Mz_k + Mz_{k-1}) + Mz_{k-1} \\
z_{k+1} &= - \omega_{k+1}(\alpha_k (I - M^\inv N) z_k - z_k + z_{k-1}) + z_{k-1} \\
&= -\omega_{k+1} \alpha_k (I - B) z_k + \omega_{k+1} z_k + (1- \omega_{k+1})z_{k-1}
\end{align*}
we can see that each $z_{k+1}$ is a linear combination of previous $z_k$ up to some coefficients, as well as $(I-B)z_k$ with some coefficients. If we know that our $z_k$ and $z_{k-1}$ terms are also formed via this recurrence relation, we can imagine that $z_{k+1}$ is probably just a polynomial in $I - B$ multiplied by $z_0$. If we allow this polynomial to change on each iteration, then this statement becomes entirely uncontroversial, since we are essentially saying \squote{we derived this from some polynomial technique, we don't really care which one}, re-ambiguating the disambiguification we did when we derived Chebyshev semi-iteration. Accordingly, we expect the order of the polynomial to be $k+1$ since $z_1 = Q_1(K) z_0$ should not be directly proportional to $z_0$, and can be seen following the rule $z_{k+1} = M^\inv r_{k+1} = z_k - \alpha_k M^\inv A p_k$ in projective conjugate gradient.

What will be useful for us however, is that to say $z_{k+1} = Q_{k+1}(I - B) z_0$ for some $k+1$ order polynomial $Q_{k+1}(t)$ is to deduce that this is equal to a different polynomial $Q_{k+1}(I - B) z_0 = (I - B P_k(B)) z_0$. This also shouldn't be too surprising to us either since each $(I - B)^a$ in $Q_{k+1}(I - B)$ with $a \le k$ will have a standard combinatorial expansion \begin{gather*}
(I - B)^a = I - a IB + ... + a I(-B)^{a-1} + (-B)^a 
\end{gather*}
so grouping all terms after $I$ just leads to $P_k(t)$ being some other arbitrary polynomial, which we were intending it to be anyway. The only thing here we need to enforce when swapping $Q_{k+1}(I - B) z_0 = (I - B P_k(B)) z_0$ is that $Q_{k+1}$ has zeroth order coefficient of one, which once again by the iteration in projective conjugate gradient, should be unsurprising since $z_{k+1} = z_k - \alpha_k M^\inv A p_k$ specializes in $k=0$ to $z_1 = z_0 - $ demonstrating the coefficient on $z_0$ is indeed one.

Then Meurant gives us not only the polynomial, he also tells us how this polynomial iterates $x_{k+1}$. \begin{label proposition}{0426PCGFirstFormAsIter}
Let $\alpha_k$ and $\omega_k$ be defined as in the preconditioned conjugate gradient method. Let $P_k(t)$ be a polynomial defined \begin{align*}
P_0(t) &= \alpha_0 \\
P_1(t) &= \omega_2 ( \alpha_0 + \alpha_1  - \alpha_0 \alpha_1 t) \\
P_{k}(t) &= \alpha_k \omega_{k+1} + \omega_{k+1} (1 - \alpha_k t) P_{k-1}(t)  - (\omega_{k+1} - 1)P_{k-2}(t)   
\end{align*}
Then the iterates $x_k$ and $z_k = M^\inv r_k = M^\inv(b - Ax_k)$  as in the preconditioned conjugate gradient method satisfy the following: \begin{enumerate}
\item The iteration for $z_k$ can be expressed $z_{k+1} = (I - B P_k(B)) z_0$.
\item The iteration for $x_k$ can be expressed $x_{k+1} = x_0 + P_k(B) z_0$.
\end{enumerate}
\end{label proposition}

\begin{my proof}{m}
Begin with part a. and proceed by induction. Recall our earlier formula for $z_k$ iteration \begin{gather*}
z_{k+1} = z_{k-1} - \omega_{k+1}(\alpha_{k} M^\inv A z_{k} - z_{k} + z_{k-1}).
\end{gather*}
where we have $B = M^\inv A$.

Consider first the base case; we recall that preconditioned conjugate gradient initializes $z_0  = M^\inv (b - Ax_0)$ and recall we set $x_{-1} = x_0$. This implies that $z_{-1} = z_0$ and so the above expression in $k=0$ with intialization $\omega_1 = 1$ simplifies to \begin{align*}
z_{1} &= z_{0} - \omega_{1}\alpha_{0} M^\inv A z_{0} \\
&= (I - \alpha_0 B) z_0 \\
&= (I - BP_0(B)) z_0
\end{align*}
as we wanted. In $z_2$ we see our formula is \begin{align*}
z_{2} &= z_{0} - \omega_{2}(\alpha_{1} B z_{1} - z_{1} + z_{0}) \\
&= z_0 - \omega_2 \alpha_1 z_1 + \omega_2 \alpha_1 z_1 - \omega_2 \alpha_1 z_0\\
&= -\omega_2 \alpha_1 (B - I))z_1 + (1 - \omega_2 \alpha_1) z_0 \\
&=  - \omega_2 \alpha_1 (B - I)(I - \alpha_0 B) z_0 + (1 - \omega_2 \alpha_1) z_0 \\
&= - \omega_2 \alpha_1 (B - \alpha_0 B^2 - I + \alpha_0 B) z_0 + (1 - \omega_2 \alpha_1) I z_0 \\
&= (I - \omega_2 \alpha_1 (\alpha_0 B + B - \alpha_0 B^2)) z_0 \\
&= (I - B \omega_2  (\alpha_0 \alpha_1 + \alpha_1 - \alpha_0 \alpha_1 B)) z_0 \\
&= (I - B P_1 (B) ) z_0 
\end{align*}
Then to prove the inductive case goes as follows. Starting from the standard iteration and applying the inductive hypothesis, \begin{align*}
z_{k+1} &= z_{k-1} - \omega_{k+1}(\alpha_{k} B z_{k} - z_{k} + z_{k-1}) \\
&= - \omega_{k+1}(\alpha_k B - I) z_k + (1 - \omega_{k+1}) z_{k-1} \\
&= - \omega_{k+1}(\alpha_k B - I) (I - BP_{k-1}(B))z_0 \\
& \hspace{3em} + (1 - \omega_{k+1}) (I - B P_{k-2}(B))z_0 \\
&= - \omega_{k+1}(\alpha_k B - I - BP_{k-1}(B) + \alpha_k B^2 P_{k-1}(B))z_0 \\
& \hspace{3em} + (1 - \omega_{k+1}) (I - B P_{k-2}(B))z_0 \\
&= Iz_0 - \omega_{k+1}(\alpha_k B - BP_{k-1}(B) + \alpha_k B^2 P_{k-1}(B))z_0 \\
& \hspace{3em} - (\omega_{k+1} - 1)  B P_{k-2}(B)z_0 \\
&= Iz_0 - B \begin{pmatrix} \omega_{k+1}\alpha_k + \omega_{k+1} (I - \alpha_k B)P_{k-1}(B) \\ - (\omega_{k+1} - 1) P_{k-2}(B)\end{pmatrix} z_0 \\
&= (I - B P_k(B)) z_0
\end{align*}

We move to part b. now. Once again we will prove by induction; since we set $x_{-1} = x_0$, the preconditioned conjugate gradient iteration \begin{gather*}
x_{k+1} = x_{k-1} + \omega_{k+1} (\alpha_k z_k + x_k - x_{k-1})
\end{gather*}
with $k=0$ is \begin{align*}
x_{1} &= x_{0} + \omega_{1} \alpha_0 z_0 \\
&= x_{0} + \alpha_0 z_0
\end{align*}
so with $P_0(B) = \alpha_0$ the base case is fulfilled. Applying $x_k = x_0 + P_{k-1}(B)z_0$ to an inductive step, we see \begin{align*}
x_{k+1} &=  \omega_{k+1} (\alpha_k z_k + x_k - x_{k-1}) + x_{k-1} \\
&=  \omega_{k+1} \alpha_k (I - B P_{k-1}(B))z_0 + \omega_{k+1}(x_0 + P_{k-1}(B)z_0) \\
& \hspace{3em} - \omega_{k+1}(x_0 + P_{k-2}(B)z_0)) + x_0 + P_{k-2}(B)z_0 \\
&=  \omega_{k+1} \alpha_k (I - B P_{k-1}(B))z_0  + x_0 + P_{k-2}(B)z_0 \\
& \hspace{3em}  + \omega_{k+1}( P_{k-1}(B) - P_{k-2}(B) )z_0 \\
&=  x_0 + \begin{pmatrix} \omega_{k+1}\alpha_k I  + \omega_{k+1} (I -B) P_{k-1}(B) \\ \quad - (\omega_{k+1} - 1) P_{k-2}(B)\end{pmatrix} z_0
\end{align*}
The terms in the brackets are exactly the recurrence relation for $P_{k}(t)$ in terms of the matrix parameter $B$, so this is \begin{gather*}
x_{k+1} = x_0 + P_k(B) z_0
\end{gather*}
\end{my proof}

This format formalizes the class of technique for which preconditioned conjugate gradient is optimal. In some sense we will say that the polynomial sequence $(P_k(t))_{k \in \nats}$ that defines the preconditioned conjugate gradient method is exactly the sequence that satisfies \begin{gather*}
\min_{(P_k(t))_{k \in \nats}} E(x_0 + P_k(t) z_0)
\end{gather*}
at each $k$, although this notation is by analogy to our earlier statement and thus not certain to be meaningful.

\begin{theorem}[Polynomial Optimality of Preconditioned Conjugate Gradient]
\emph{(Stated as in Meurant)}

Consider all the iterative methods that can be written as \begin{gather*}
\bar x_{k+1} = \bar x_0 + Q_k(B) \bar z_0,\\
\bar x_0 = x_0 \\
M \bar z_0 = b - A \bar x_0
\end{gather*}
where $Q_k$ is a $k$'th degree polynomial. Of all these methods, \texttt{PCG} is the one that minimizes $E(\bar x_k)$ at each iteration.
\end{theorem}

Meurant proves this fairly explicitly over the course of three pages so unless it seems pertinant to update this page, I will merely refer to it as theorem 6.11 on page 180 of his book, lest I rewrite it as a six page proof. I must note however that at the stage he gives the proof, he defines $E(x)$ not in the way we have done so far, but as \begin{gather*}
E(z) = (z - x)^T A (z - x) 	 
\end{gather*}
the square of our definition, but with $x$ used to denote the solution to $Ax = b$ rather than $x_*$ as we have done so far.

Nonetheless this theorem is remarkable; what it tells us is that if we want to come up with a method that could ever beat preconditioned conjugate gradient, it would need to be one that can never be expressed in terms of a polynomial iteration in its preconditioned residuals. Failing that (or otherwise, taking some method and showing that it can be expressed in the terms above), we know that preconditioned conjugate gradient will simply beat other methods on convergence rate.

In a manner of speaking, this tells us that without additional assumptions, preconditioned conjugate gradient and its mathematically equivalent algorithms are often just the best we can get. That said, we will soon see that our deeper knowledge of these algorithms is rewarded when additional assumptions about $A$ are available.

\subsection{Vital Takeaways}

It is at this point that I must admit this article has gone on much longer than I had initially intended. In part this is because I began this with intention to help me order my thoughts as I continued reading, but it has now become longer than makes sense for the structure I had given it. We aim to conclude now as soon as possible, but first we must discuss some matters which only appeared to me once I was I had become much more familiar with this literature.

This section will serve as a \squote{skip to the juicy bits} tl;dr. Where notions such as preconditioning in abstract and conjugate gradient were very quickly mentioned to me upon taking interest in the field (although taking significant time to fully understand), it took much more time before someone mentioned \dquote{GMRES}, or I encountered the Joly-Meurant method. Or until I came to understand why the field of iterative methods so liberally spoke of solving $Mz = r$ systems as though this were an easier problem than the $Ax=b$ problem we initially aimed to solve.

These topics we will sound off here before concluding.

\subsubsection{Convergence Rate of Conjugate Gradient}

When I began this project, I was under the impression that conjugate gradient methods could not be expected to give a good approximation of of a solution in any less than $n$ iterations since it was described to me as a method which ostensibly ran Gram-Schmidt. While in a manner of speaking, we have come to see that the latter is true, the former turns out not to be. This could perhaps be obvious to us already, on account of the fact we showed that under conjugate gradient, our residual vectors end up orthogonal to one another. In fact it is not that $n$ iterations are required, but that in $n$ iterations and in exact arithmetic, an exact solution is found.

We are given the following error bound by Meurant in chapter 6.4 (specifically theorem 6.18) of his book. \begin{theorem}[Convergence Rate of Conjugate Gradient]
Let $x_0$ be an initial guess to solve $Ax = b$ and $x_k$ a sequence of guesses as iterated by conjugate gradient. Then with \begin{gather*}
E(z) = (z - x_*)^T A (z - x_*)
\end{gather*}
(this is not the definition of $E$ we used so far) we have the bound \begin{gather*}
E(x_k) \le 4 \left( \frac{\sqrt{\kappa} - 1}{\sqrt{\kappa} + 1} \right)^{2k} E(x_0)
\end{gather*}
where $\kappa = \lambda_{\text{max}}/\lambda_{\text{min}}$ the ratio of the largest eigenvalue of $M^\inv A$ its smallest eigenvalue.
\end{theorem}

This tells us already that there exists some constant for $M^\inv A$ for which we can expect the error in iterations to decline exponentially. We see that this exponent is closest to zero when the eigenvalues of $M^\inv A$ are close together, hence the power of preconditioning. One could imagine then that it is possible to find oneself in a situation with a particularly poorly conditioned matrix for which this constant is not very small, and to this, Meurant says something peculiar.

\emph{"It has been observed experimentally that CG exhibits what may be called a superlinear rate of convergence, that is to say, the convergence becomes faster and faster as CG proceeds, see Figure 6.1. The explanation of this phenomenon is that the components of the error on the eigenvectors corresponding to the extreme eigenvalues are rapidly dampened during the iterations and then, the convergence is faster because the “effective” condition number gets smaller."}

\figurepngwithcaption{./MeurantsIterfigure6-1}{The figure 6.1 referenced by Meurant, originally captioned "log10 of the maximum norm of the error with CG for the Poisson problem on a 20 × 20 mesh". By this he means conjugate gradient was run without a preconditioner. Note that in exact arithmetic, the algorithm would have converged by iteration 20. Due to a combination of poor conditioning and rounding error, it takes significantly more iterations, but eventually hits a point of order-of-magnitude improvements per ten iterations.}

Informally, it may be reasoned that what Meurant means here is to do with the fact that Krylov subspaces are successively extended in conjugate gradient. That is, in the same way that one simple algorithm one might run to extract the largest eigenvalue of a matrix would be to apply it to a vector over and over and over and see how much it grows since this should bias towards eigenvectors with larger eigenvalues, it may be that each iteration of conjugate gradient subtracts the contribution of more dominant eigenvectors in $x_* - x_k$. If true, this reasoning would suggest that $x_* - x_k$ for large $k$ simply has very little contribution from its large eigenvalue components, and so on each iteration there is a sense in which it is not $\lambda_{\text{max}}$ that is most appropriate for $\kappa$ but a smaller eigenvalue.

As we will see in our subsection on Joly-Meurant, this insight is of critical importance to rapid convergence in contexts where we have many $Ax = b$ problems that share an $A$ but may change in $b$.


\subsubsection{Generalized Minimal Residual Method (GMRES) and Methods for Nonsymmetric $A$}

It would be criminal of me not to speak on GMRES before concluding, although I must admit I devoted little of my time to studying the method. The principle circumstance in which one might use GMRES is when one has a non-symmetric $A$, and when that is the case, it is of course possible to solve the $A^T A x = A^T b$ problem which has the same solution, with only a matrix-matrix multiplication and a matrix-vector multiplication up front.

Nonetheless, as conjugate gradient is an orthogonal projection method, GMRES is an oblique projection method using $\mathcal L = A \mathcal K$ that follows from full orthogonalization method and Arnoldi's method.

The method calls for running some version of Arnoldi's method in the way we described before (although probably with appropriate rearrangements to deal with the instabilities in naive Gram-Schmidt) to obtain an orthonormal basis $V$ in $\mathcal K_m(A,r_0)$ followed by using our previously mentioned equation $AV = \tilde V \tilde H$ (these were $\tilde V$ which is $V$ along with the extra basis vector $v_{m+1}$ derived at the end of the algorithm and $\tilde H$ the Gram-Schmidt coefficient matrix with the first column omitted) to say the following. If the solution $x_*$ has $x_* = x_0 + \delta$ where $\delta \in \mathcal K_m(A,r_0)$ then \begin{align*}
b - Ax &= b - A(x_0 + V \vec c)  \\
&= (b - Ax_0) - AV \vec c \\
&= r_0 - \tilde V \tilde H \vec c \\
&= H_{11} v_1 - \tilde V \tilde H \vec c \\
&= \tilde V (H_{11} e_1 - \tilde H \vec c)
\end{align*}
Failing $x_0 + \delta = x_*$, GMRES says to solve for $\vec c$ which minimizes $\lVert H_{11} e_1 - \tilde H \vec c \rVert $ via least squares.

While many sources have assured me that this is a circumstance in which least squares runs especially efficiently,  part of my disinterest in this algorithm is that I am yet to find any description that cleanly convinces me of that, much less provides any of the radical performance guarentees we have for preconditioned conjugate gradient on $A^T A x = A^T b$.

\subsubsection{Solving many $Ax=b$ With Same $A$: The Joly-Meurant Method}

At the beginning of this article I mentioned that my intention with all of this was to solve the kinds of $Ax=b$ problems one derived from the weak formulation of a partial differential equation in the course of a finite element method. Such applications, among others such as finite difference, finite volume, etc. often call for applying a difference method in the time variable, producing a large quantity of $Ax = b$ problems where $b$ encodes information related to the previous timestep and $A$ encodes the fundamental dynamics between timesteps, unchanging.

Something interesting about the conjugate gradient method one may notice during its derivation is that since it is based on the direct Lanczos method where we required our $p_k$ basis were $A$-orthogonal, one might deduce that changing $b$ but not changing $A$ preserves this property, allowing the $P$ basis to be reused. This intuition is formalized in what I have decided to call the \hyperlink{https://gerard-meurant.fr/cgc_1992.pdf}{Joly-Meurant method}, which appears to have been developed for complex $Ax=b$ problems.

Joly-Meurant effects a framework in the same way that splitting methods provide a calculation one could to to iterate without telling you how to split $A = M - N$, or how a projection method tells you what you have to do to find your update $\delta = V \vec c$ but not how to choose $\mathcal K$ or $\mathcal L$. Their iteration, for a given $x_0$ and two matrices $H$ symmetric positive definite and $K$ satisfying $(K+K^T)$ being either strictly positive definite or negative definite, and with $N = A^T H A$, $g_0 = A^T H r_0$ and $p_0 = K g_0$, looks like the following: \begin{gather*}
\alpha_k = \frac{g_k \cdot p_k}{p_k \cdot N p_k} \\
x_{k+1} = x_k + \alpha_k p_k \\
r_{k+1} = r_k - \alpha_k A p_k \\
g_{k+1} = g_k - \alpha_k N p_k \\
\beta_{k+1}^l = - \frac{(K g_{k+1}) \cdot (N p_l)}{p_l \cdot N p_l}, \hspace{2em} 0 \le l \le k \\
p_{k+1} = K g_{k+1} + \sum_{l = 0}^k \beta_{k+1}^l p^l 
\end{gather*}

This should look suspiciously like conjugate, and that's because it is in a general sense. If we set $H = A$ with $A$ symmetric positive definite and set $K = I$, we get $N = A^T = A$ and our $g_k$ vectors become exactly equal to $p_k$, removing the need for the triangle of $\beta_k^l$ values. With $H = I$ and $K = A^{-T}$ one derives another algorithm known as conjugate residual, and in general many other conjugate methods can be derived through choices of $H$ and $K$.

What is remarkable about this framework is that, independent of a choice of $H$ and $K$, if one is to change $b$ but keep the vectors $p_k$, many of the orthogonality guarentees one would want from the newly generated $g_k^i$ generated from $g_0^i = A^T A r_0^i = A^T A (b^i - Ax_0^i)$ solved at $i$'th $b$ remain true. That is, just as in conjugate gradient we have $r_i \cdot r_j = 0 $ and $p_i \cdot A p_j = 0$ for all $i \neq j$, in Joly-Meurant we have \begin{align*}
p_k \cdot N p_j = 0, \hspace{2em} &k \neq l \\
g_k \cdot p_j = 0, \hspace{2em} &j < k \\
g_j \cdot p_k = g_0 \cdot p_k, \hspace{2em} &j \le k \\
g_k \cdot p_k = g_k \cdot K g_k \hspace{2em} & \\
(K g_k) \cdot (Np_k) = p_k \cdot N p_k \hspace{2em} & \\
g_k \cdot K g_j = 0, \hspace{2em} & j < k 
\end{align*}
with $g_k$ referring to the \squote{seed system}, an initial $Ax = b^0$ we use to calculate $p_k$ and in this case $g_k$, we find that on successive systems we still have \begin{align*}
g_k^i \cdot p_j = 0, \hspace{2em} & j < k \\
g_k^i \cdot p_j = g_0^i \cdot p_k , \hspace{2em} & j \le k\\
g_k^i \cdot p_k = g_k^i \cdot K g_k, \hspace{2em} &  g_k := g_k^0
\end{align*}
which according to Meurant is enough for the algorithm to keep working. My understanding of precisely why, in terms more precise than what is given above, is unfortunately poor, however I am elated at the consequence.

This brings us to the radical idea that a conjugate gradient method can work with zero matrix multiplications per iteration once the seed system is solved. The particular method proposed by Joly and Meurant here seems to be under the expectation that many $Ax = b^i$ systems are solved in parallel with one merely chosen as the seed system, however here it is written as a method to be run one $b^i$ after another specifically in the context of conjugate gradient.

The Joly-Meurant Complex Conjugate Gradient algorithm can be thought of as the following: \begin{itemize}
\item Choose $x_0 = x_0^0$ and initialize all other memory required to run preconditioned conjugate gradient. Set $s = 0$ as the system designated as seed-system
\item Run preconditioned conjugate gradient, storing $p_k$ vectors and caching $A p_k$ vector and $p_k \cdot A p_k$ dot product results. Store the number of iterations the seed system needs to converge to satisfactory bound as integer $m$.
\item For each $b^i$, set $g_0^i = A^T A  r_0^i$ (setting $x_0^i$ as the solution calculated for system $i-1$ if $b^i$ are related by some time-iteration sequence such as in finite element or finite difference) solve $Ax^i = b^i$ by these steps for each iteration $k=0,\dots $ until convergence, using cached values: \begin{enumerate}
\item Set \begin{gather*}
\alpha_k^i = \frac{g_k^i \cdot p_k }{p_k \cdot A p_k}
\end{gather*}
\item Set \begin{gather*}
x_{k+1}^i = x_k^i + \alpha_k^i p_k
\end{gather*}
\item Set \begin{gather*}
r_{k+1} = r_k - \alpha_k^i A p_k
\end{gather*}
\item Set \begin{gather*}
g_{k+1}^i = g_k^i - \alpha_k^i A p_k 
\end{gather*}
\item If $k+1 > m$, check next step. Else, if $\lVert r_{k+1}^i \rVert$ is satisfactory, terminate loop in $k$ and move to next $i$.
\end{enumerate} 
\item If in the previous step a system requires more than $m$ iterations (or indeed if running preconditioned conjugate gradient and these other systems in parallel, and the seed system converges before some of the others) then perform the following steps to transfer ownership of the seed system title/property to another system and resume generating new $p_k$ vectors: \begin{enumerate}
\item Set $\sigma = s$ to denote the old seed system.
\item Choose new $s$ as the system $i$ with largest $\lVert r_{k+1}^i \rVert$
\item Set \begin{gather*}
\beta_{k+1} = \frac{g_{k+1}^s \cdot g_{k+1}^\sigma }{p_k \cdot A p_k}
\end{gather*}
\item Set \begin{gather*}
p_{k+1} = g_{k+1}^s + \beta_{k+1} p_k
\end{gather*}
\item Resume step 2's preconditioned conjugate gradient with the new seed system, caching $p_k$, $Ap_k$ and $p_k \cdot A p_k$ results.
\end{enumerate}
\end{itemize}

Note in particular that if vectors and inner products are cached as described above, the sum total calculations in non-seed-system iterations is one dot product in $g_k^i \cdot p_k$, three two scalar multiplications $\alpha_k^i p_k$ and $\alpha_k^i A p_k$, and three additions, and an additional $r_{k+1}^i \cdot r_{k+1}^i$ product computed only as desired to check for convergence. Counting all but the dot products as trivial, this amounts to maybe 1.1 dot products per iteration, assuming convergence is only checked every ten times.

Moreover, if the step specified in which the solution to the previous $i-1$ system is used as the initial guess for the next system, by the argument made earlier, it is likely that the system initializes with much smaller \squote{effective condition number}, meaning the number of necessary iterations after the first is much smaller.

In general, I find myself somewhat shocked that there is not more awareness of this algorithm. 

\subsubsection{Symmetric Preconditioners and Efficient $Mz = r$ Solution}

Unfortunately I did not ultimately read as much about preconditioners as I had initially set out to do, however with respect to my purposes, I have certainly learned all I needed to. Meurant notes about diagonal preconditioners \emph{"The simplest ideas (which unfortunately are by far not the most efficient ones) are based on the classical iterative methods"} so we will put diagonal preconditioners out of our minds as a topic already studied. 

Throughout my review, I was frequently confused as to why methods would include a step that says \squote{solve $M z = r$} as though this would be any easier than solving the base $Ax = b$ problem we were aiming for in the first place. The answer to that, in a manner of speaking, is that I had never fully internalized the usefulness of $LU$-decompositions. A system $Lx = b$ with $L$ triangular is not merely \squote{easy to solve}, it violates the regime of thinking in which one considers finding $x$ to being equivalent to calculating $L^\inv b$. This was the lesson we learned in Gauss-Seidel splitting, when we saw that $A = M - N$ with $M = D + L$ with $L$ all elements below the diagonal in $A = D + U + L$, had \begin{gather*}
(D + L) x_{k+1} = b - Ux_k 
\end{gather*}
easy to solve since we merely use previously calculated $x_{k+1}^i$ elements of the $x_{k+1}$ vector, actually computing \begin{gather*}
x_{k+1}^i = \frac{1}{A_{ii}} \left( b^i - \sum_{i < j} A_{ij} x_k^j - \sum_{i > j} A_{ij} x_{k+1}^j \right).
\end{gather*}

The thing is that nothing stops us from doing this trick in any order that we like. We had to use lower diagonals only if we calculated each index $i$ of $x$ starting from $i=1$ and going up to $n$, but we could do this in reverse.

This is especially useful to us since preconditioned conjugate calls for a symmetric preconditioner, yet never actually needs $M^\inv$, only solutions to $Mz = r$ systems. Meurant highlights the SSOR preconditioner (symmetric successive over relaxation preconditioner) which for a symmetric $A = D + L + L^T$ has these properties as well as a tuning paramter $0 < \omega < 2$. \begin{gather*}
M = \frac{1}{\omega(2 - \omega)} (D + \omega L)D^\inv (D + \omega L^T)
\end{gather*}

When solving the $Mz = r$ step, this amounts to merely two additional matrix multiplications in a similar manner to the trick we learned in Gauss-Seidel, applied twice. First we calculate an intermediaery $z_*$ which solves $(D + \omega L^T)z_* = r$ by \begin{gather*}
z_*^i = \frac{1}{A_{ii}} \left( r^i - \sum_{i < j} A_{ij} z_*^i \right),
\end{gather*}
we can multiply by our coefficient at the same time as we multiply through our diagonal $(\omega(2-\omega))^\inv D^\inv z_{**} = z_*$ in the obvious way, followed by the final step \begin{gather*}
z^i = \frac{1}{A_{ii}} \left( z_{**}^i - \sum_{i > j} A_{ij} z^i \right),
\end{gather*}
allowing the trick in Gauss-Seidel to be applied to the upper triangle of $A$ followed by the lower triangle.

Now, to apply this, one must choose $\omega$. In the section Meurant speaks of this as an explicit preconditioner, he gives little detail as to how to select $\omega$, but in his earlier discussion of SSOR as a splitting method, he mentions prior work by Young (the inventor of the method) showing that \begin{gather*}
\omega = \frac{2}{1 + \sqrt{2(1- \rho(-D^\inv(L + U))^2)}}
\end{gather*}
is a \squote{good} choice. Of course this refers to $\rho$ as the spectral radius, giving the largest eigenvalue of a matrix, which is not easily computed except by heuristic or approximation, but if you have ten or twenty matrix multiplications to spare then you are likely to get a good approximation via \begin{gather*}
\lambda_{\text{max}} \approx \left({\frac{x \cdot A^k x}{x \cdot x}} \right)^{1/k}
\end{gather*}
for some large $k$; this technique is helped by using a few different $x$ values, which can be done on a GPU, and will converge on $\lambda_{\text{max}}$ more quickly the more ill conditioned the starting matrix is.

\subsection{Conclusions and Future Directions}

I can't say I knew what I was going to write when I began this article, but it is clear that the majority of it became focused on understanding why exactly conjugate gradient methods work. Using that understanding, we are able to understand how to improve it with preconditioners without $M^\inv A$'s failure of symmetry becoming a problem even if $M$ and $A$ are symmetric, and we are able to understand how to cache intermediate quantities so that successive $Ax = b$ problems are easier to solve.

The outline of lessons learned to the functioning of the conjugate gradient method are as follows. We first understand that a projection technique in which an update vector $\delta$ in a subspace is used to define an iteration $x_1 = x_0 + \delta$ where the new residual $b - Ax_1$ is expected to be orthogonal to that same subspace constitutes a Gram-Schmidt method as well as an inversion in the coefficient matrix calculated during Gram-Schmidt, the $\delta = V\vec c$ where $\vec c = H^\inv V^T r$ step. Next, we understand that by omitting the first column and last row of the coefficient matrix $H$, having chosen the subspace to be a Krylov subspace means that symmetry in $A$ leads to symmetry in $H$, making it tridiagonal. Knowing this, we also find that there exist iteratively defined bidiagonal $LU$ decompositions in tridiagonal matrices, as well as iteratively defined inverses for these  bidiagonal $L$ and $U$, leading to a reformulation of the full orthogonalization method of $x_{k+1} = x_k + V \vec c = x_0 + V(H^\inv V^T r_k)$ as $x_{k+1} = x_k + (V U^\inv) (L^\inv e_k) = x_k + P z_k$. Since these iterative definitions allow us to express the $x_{k+1} = x_k + Pz_k$ update as single element updates in $z_k$, we can express large single projection methods as iterative methods once again, making the Krylov subspace $\mathrm{span}(r_0,Ar_0,..., A^n r_0)$ without complication. Finally, since each $\mathrm{span}(r_0,Ar_0,..., A^{k+1} r_0)$ defines a unique new direction with respect to $\mathrm{span}(r_0,Ar_0,..., A^k r_0)$, the new update directions are always uniquely defined, and free to be calculated by much more efficient much more stable means than a highly complicated Gram-Schmidt. Seemingly for all the reasons that repeated multiplication of a matrix on a random vector can result in an estimate of the largest eigenvalue of the matrix, the use of Krylov subspaces means that conjugate gradient preferentially subtracts the contribution of eigenvectors associated to larger eigenvalues in the residual, resulting in rapid convergence.

As a consequence of these lessons and the lessons required to build up to it, we also learn the following vital lessons: \begin{itemize}
\item Via Joly-Meurant, future or simultaneous $Ax = b'$ problems which share a $A$ as the original matrix but have $b'$ instead of $b$ may reuse the $A$-orthogonal $p_k$ basis, resulting in a method which only requires a single dot product and zero matrix multiplications on each iteration.
\item Preconditioning and the associated $Mz = r$ problems which must be frequently solved to use preconditioning are trivial, tantamount to an additional matrix multiplication, so long as $M$ is chosen as a product of triangular matrices. This is because forwards and back substitution may be performed at similar speeds as matrix multiplication, but in particular, since we may construct $M$ without a multiplicative $LU$-decomposition, we obtain effective preconditioners $M$ with roughly the same sparsity as $A$.
\end{itemize}

Finally, I'll end with what I didn't discuss but would have liked to. That is, no consideration is made in this article about the actual runtime of these algorithms on hardware, or the role of parallelisation in their optimization. One must remember that the primary sources in this review were written twenty five years ago, in a very different computing landscape. Since we conclude that Joly-Meurant is a dot-product workload, I find it highly likely that GPU acceleration is just as often harmful as it is helpful by contrast to many core CPUs methods. In fact, most of the operations described towards the end here are likely to suffer primarily from memory bandwidth limitations rather than arithmetic cost. A future article may explore the runtime differences between different implementations. Additionally, I was unable to find satisfactory graphs for the convergence rate of conjugate gradient and preconditioned conjugate gradient in larger systems, and that is definitely something I would like to visualize. Algorithms also made frequent mention of instability due to rounding error, and it became increasingly frustrating that I could not characterize this instability or how significant it was at higher precisions; I have some reason to suspect that (especially for poorly conditioned matrices) dot product steps with their large floating point additions were the principle drivers of rounding errors since smaller floating point numbers are rounded to fit with the largest value in the $x \cdot y = \sum_{i} x^ i y^i$ sum.

It is worth noting that both book authors, Yousef Saad and Gerard Meurant, are still academically active and contributing to the study of the discussed topics in our changing computing landscape. While Meurant has increasingly become focused on historical matters of mathematics, Saad publishes something on parallelization in linear solving methods every few years for various applications. In particular he seems interested in the Schur-compliment and its applications in paralellizing conjugate gradient methods by approaching $A$ as a block matrix.

\subsection{References and Further Reading}

The following two books formed the backbone of the above narration: \begin{itemize}
\item Yousef Saad, \emph{Iterative methods for sparse linear systems.} Society for Industrial and Applied Mathematics, 2003.
\item Gerard Meurant, \emph{Computer solution of large linear systems.} Vol. 28. Elsevier, 1999. (I used the 2024 online pdf edition)
\end{itemize}

Additional papers referenced are as follows: \begin{itemize}
\item Emrah Kılıç and Pantelimon Stanica. \emph{"The inverse of banded matrices."} Journal of Computational and Applied Mathematics 237.1 (2013): 126-135. (Used for $LU$-decomposition of tridiagonal $H$ and inversion of bidiagonal $L$ and $U$.)
\item Paul Concus, Gene H. Golub, and Dianne P. O'Leary. \emph{"A generalized conjugate gradient method for the numerical solution of elliptic partial differential equations."} Sparse matrix computations. Academic Press, 1976. 309-332. (Gave early formulations of preconditioned conjugate gradient methods.)
\item David M. Young. \emph{Iterative Solution of Large Linear Systems}. Academic Press, 1971 (Cited by Meurant as discussing SOR and SSOR methods, as well as providing heuristics for choosing optimum $\omega$ in successive over relaxation methods.)
\item Germund Dahlquist and Åke Björck. \emph{Numerical methods in scientific computing, volume I}. Society for Industrial and Applied Mathematics, 2008. (Meurant cites this text for the properties of Chebyshev polynomials given here.)
\item Pascal Joly, and Gérard Meurant. \emph{"Complex conjugate gradient methods."} Numerical Algorithms 4.3 (1993): 379-406. (Gives the Joly-Meurant method)
\item Pascal Joly. \emph{"Résolution de systemes linéaires avec plusieurs seconds membres par la méthode du gradient conjugué."} Tech. ReportTech. ReportR-91012, PublicationsduLaboratoired'Analyse Numerique, Universite Pierre et Marie Curie, Paris (1991). (Unfortunately I was never able to find a copy of this, and in any case it is unlikely a translated version exists. Meurant cites this for the proof that $g_k^i$ and $p_k$ vectors maintain the necessary orthogonality conditions required for Joly-Meurant to keep working when $p_k$ are derived only for $b^0$.)
\end{itemize}
\end{document}